\input{cpp-macros}

\setchapterpreamble[u]{\margintoc}
\chapter{Eliminating Reflection from Type Theory}

\todo{Taken from the CPP article (for now as is).}

\todo{Cite paper and authors}

% \begin{abstract}
  Type theories with equality reflection, such as extensional type
  theory (ETT), are convenient theories in which to formalise
  mathematics, as they make it possible to consider provably equal terms
  as convertible. Although type-checking is undecidable in this context,
  variants of ETT have been implemented, for example in \NuPRL and more
  recently in \Andromeda.
  %
  The actual objects that can be checked are not proof-terms, but
  derivations of proof-terms. This suggests that any derivation of ETT
  can be translated into a typecheckable proof term of intensional type
  theory (ITT).
%
  However, this result, investigated categorically by
  Hofmann in 1995, and 10 years later more
  syntactically by Oury, has never given rise to
  an effective translation.
  %
  In this paper, we provide the first effective syntactical translation
  from ETT to ITT with uniqueness of identity proofs and functional
  extensionality. This translation has been defined and proven correct
  in \Coq and yields an executable plugin that translates a derivation
  in ETT into an actual \Coq typing judgment.  Additionally, we show how
  this result is extended in the context of homotopy type theory to a
  two-level type theory.
% \end{abstract}

\section{Introduction}

Type theories with equality reflection, such as extensional type
theory (ETT), are convenient theories in which to formalise
mathematics, as they make it possible to consider provably equal terms as
convertible, as expressed in the following typing rule:
%
\begin{equation}
  \label{eq:reflection}
  \infer[]
    {\xisterm{\Ga}{e}{\Eq{A}{u}{v}}}
    {\xeqterm{\Ga}{u}{v}{A}}
  %
\end{equation}
%
Here, the type $\Eq{A}{u}{v}$ is Martin-Löf's identity type with only
one constructor $\refl{} u : \Eq{A}{u}{u}$ which represents
proofs of equality inside type theory, whereas $u \equiv v : A$ means
that $u$ and $v$ are convertible in the theory---and can thus be
silently replaced by one another in any term.
%
Several variants of ETT have been considered and implemented, for
example in \NuPRL\sidenote{Although the reflection rule is provable in \NuPRL,
its calculus is based on realisability rather than on intensional type theory
plus reflection.}~\cite{DBLP:conf/cade/AllenCEKL00} and more recently in
\Andromeda~\cite{andromeda}.
%
The prototypical example of the use of equality reflection is the
definition of a coercion function between two types $A$ and $B$ that
are equal (but not convertible) by taking a term of type $A$ and
simply returning it as a term of type $B$:
\[
\lambda\ A\ B \ (e:A = B)\ (x:A).\ x
: \Pi\ A\ B.\ A = B \to A \to B.
\]
%
%
In intensional type theory (ITT), this term does not type-check
because $x$ of type $A$ can not be given the type $B$ by conversion.
%
In ETT, however, equality reflection can be used to turn the witness
of equality into a proof of conversion and thus the type system
validates the fact that $x$ can be given the type $B$.
%
This means that one needs to guess equality proofs
during type-checking, because the witness of equality has been lost at
the application of the reflection rule. Guessing it was not so hard in this
example but is in general undecidable, as one can for instance encode the
halting problem of any Turing machine as an equality in ETT.
%
That is, the actual objects that can be checked in ETT are not terms,
but instead derivations of terms.
%
It thus seems natural to wonder whether any derivation of ETT can be
translated into a typecheckable term of ITT.
%
And indeed, it is well know that one can find a corresponding term of
the same type in ITT by \emph{explicitly} transporting the term $x$
of type $A$ using the elimination of internal equality on the witness
of equality $e$, noted $\translpo{e}$:
%
\[
  \lambda\ A\ B \ (e:A = B)\ (x:A).\ \translpo{e}\ x
  : \Pi\ A\ B.\ A = B \to A \to B.
\]
%
This can be seen as a way to make explicit the \emph{silent} use of
reflection.
%
Furthermore, by making the use of transport as economic as possible,
the corresponding ITT term can be seen as a compact witness of the
derivation tree of the original ETT term.

This result has first been investigated categorically in the
pioneering work of \cite{hofmann1995conservativity,HofmannPhD},
%
by showing that the term model of ITT can be turned into a model of ETT by
quotienting this model with propositional equality.
%
However, it is not clear how to extend this categorical construction
to an explicit and constructive translation from a derivation in ETT to
a term of ITT.
%
In 2005, this result has been investigated more syntactically by
\cite{oury2005extensionality}. However, his presentation does not give
rise to an effective translation.
%
By an \emph{effective} translation we mean that it is entirely
constructive and can be used to deterministically \emph{compute} the
translation of a given ETT typing derivation.
%
Two issues prevent deriving an \emph{effective} translation from Oury's
presentation, and it is the process of actual formalisation of the
result in a proof assistant that led us to these discoveries. First, his
handling of related contexts is not explicit enough, which we fix by
framing the translation using ideas coming from the parametricity
translation (Section \ref{sec:heteq}). Additionally, Oury's proof
requires an additional axiom in ITT on top of functional
extensionality (FunExt)
and uniqueness of identity proofs, that has no clear motivation and can
be avoided by considering an annotated syntax (Section
\ref{sec:ettsyntax}).

\paragraph*{Contributions.}
In this paper, we present the first effective syntactical translation
from ETT to ITT (assuming uniqueness of identity proofs (UIP) and
FunExt in ITT).
%
By syntactical translation, we mean an explicit translation from a
derivation $\xisterm{\Ga}{t}{T}$ of ETT (the $\exmark$ index testifies that it
is a derivation in ETT) to a context $\Ga'$, term
$t'$ and type $T'$ of ITT such that $\isterm{\Ga'}{t'}{T'}$ in ITT.
%
This translation enjoys the additional property that if $T$ can be
typed in ITT, \ie $\istype{\Ga}{T}$, then $T' \equiv T$.
%
This means in particular that a theorem proven in ETT but whose
statement is also valid in ITT can be automatically transferred to a theorem
of ITT. For instance, one could use a \emph{local} extension of the
\Coq proof assistant with a reflection rule, without being forced to rely on
the reflection in the entire development.

This translation can be seen as a way to build a syntactical model of
ETT from a model of ITT as described more generally
in~\cite{boulier17:next-syntac-model-type-theor} and has been entirely
programmed and
formalised in
\Coq~\cite{coq}. For this, we rely on
\TemplateCoq\sidenote{\url{https://metacoq.github.io/metacoq/}}~\cite{DBLP:conf/itp/AnandBCST18},
which provides a reifier for \Coq terms as represented in \Coq's kernel
as well as a formalisation of the type system of \Coq.
%
Thus, our formalisation of ETT is just given by adding the reflection
rule to a subset of the original type system of \Coq.
%
This allows us to extract concrete \Coq terms and types from a closed
derivation of ETT, using a little trick to incorporate Inductive types
and induction. We do not treat cumulativity of universes which is an
orthogonal feature of \Coq's type theory. It would also complicate the
proof which relies on uniqueness of typing.

\paragraph*{Outline of the Paper.}

Before going into the technical development of the translation, we
explain its main ingredients and differences with previous works.
Then, in Section~\ref{sec:syntax-features}, we define the extensional
and intensional type theories we consider. In
Section~\ref{sec:relation}, we define the main ingredient of the
translation, which is a relation between terms of ETT and terms in ITT.
%
Then, the translation is given in Section~\ref{sec:translation}.
Section~\ref{sec:form-with-templ} describes the \Coq formalisation and
Sections~\ref{sec:axioms} and \ref{sec:related-works} discuss
limitations and related work.
%
The main proofs are given in detail in Appendices~\ref{sec:proof-fund-lemma}
and~\ref{sec:corr-transl}.


The \Coq formalisation can be found in
\url{https://github.com/TheoWinterhalter/ett-to-itt}.


\subsection{On the Need for UIP and FunExt}

Our translation targets ITT plus UIP and FunExt,
which correspond to the two following axioms (where $\Ty{i}$ denotes
the universe of types at level $i$):
$$
\begin{array}{r@{~}c@{~}l}
\mathtt{UIP} & : & \Pi(A: \Ty{i})\ (x \ y:A)\ (e \ e' : x = y).\ e = e' \\
\mathtt{FunExt} & : & \Pi(A : \Ty{i})\ (B : A \to \Ty{i})\
(f \ g : \Prod{x:A} B\ x). \\
&& (\Prod{x : A} f\ x = g\ x) \to f = g
\end{array}
$$
%
The first axiom says that any two proofs of the same equality are
equal, and the other one says that two (dependent) functions are equal
whenever they are pointwise equal\sidenote{In Homotopy Type
  Theory (HoTT)~\cite{hottbook}, FunExt is
  stated in a more complete way, using the notion of adjoint
  equivalences, but this more complete way collapses to our simpler
  statement in presence of UIP.}.
%
These two axioms are perfectly valid statements of ITT and they can be
proven in ETT.
%
Indeed, UIP can be shown to be equivalent to the Streicher's axiom K
$$
\begin{array}{rcl}
\mathtt{K} & : & \Prod{A: \Ty{i}} \Prod{x :A} \Prod{e : x = x} e = \refl{x} \\
\end{array}
$$
using the elimination on the identity type. But K is provable in ETT
by considering the type
$$
\Prod{A: \Ty{i}} \Prod{x \ y:A} \Prod{e : x = y} e = \refl{x} \\
$$
which is well typed (using the reflection rule to show that $e$ has
type $x= x$) and which can be inhabited by elimination of the identity
type.
%
In the same way, FunExt is provable in ETT because
$$
\begin{array}{cll}
&\Prod{x : A} f\ x = g\ x \\
\to & x : A \vdash f\ x \equiv g\ x &
\mbox{by reflection} \\
\to &  (\lambda (x:A). {f \ x}) \equiv (\lambda{(x:A)}.{g \ x}) &
\mbox{by congruence of $\equiv$} \\
\to &  f \equiv g & \mbox{by $\eta$-law} \\
\to &  f = g
\end{array}
$$

Therefore, applying our translation to the proofs of those theorems in ETT
gives corresponding proofs of the same theorems in ITT.
%
However, UIP is independent from ITT, as first shown by Hofmann and
Streicher using the groupoid model~\cite{groupoid-interp}, which has
recently been extended in the setting of univalent type theory using
the simplicial or cubical models~\cite{kapulkin2012simplicial,coquand:cubical}.
%
Similarly, FunExt is independent from ITT,
it is folklore but has recently been formalised by Boulier \emph{et al.} using a
simple syntactical translation~\cite{boulier17:next-syntac-model-type-theor}.

Therefore, our translation provides proofs of axioms independent from
ITT, which means that the target of the translation already needs to
have both UIP and FunExt.
These last two elements are only necessary in ITT and could be removed from ETT
but this allows us to consider only one syntax for both.
%
Part of our work is to show formally that they are the only axioms
required.

\subsection{Heterogeneous Equality and the Parametricity Translation}
\label{sec:heteq}

The basic idea behind the translation from ETT to ITT is to interpret
conversion using the internal notion of equality, \ie the identity type.
%
But this means that two terms of two convertible types that were comparable
in ETT become comparable in ITT only up-to the equality between the
two types. One possible solution to this problem is to consider a
native heterogeneous equality, such as \emph{John Major
equality} introduced by \cite{mcbride2000dependently}.
%
However, to avoid adding additional axioms to ITT as done by
\cite{oury2005extensionality}, we prefer to encode this
heterogeneous equality using the following dependent sums:
$$\Heq{T}{t}{U}{u} := \Sum{p:\Eq{}{T}{U}} \Eq{}{\translpo{p}\ t}{u}.$$
%

During the translation, the same term occurring twice can be
translated in two different manners, if the corresponding typing
derivations are different. Even the types of the two different
translations may be different.
%
However, we have the strong property that any two translations of the
same term only differ in places where transports of proof of equality have been
injected.
%
To keep track of this property, we introduce the relation $t \sim t'$
between two terms of ITT, of possibly different types.
%
The crux of the proof of the translation is to guarantee that for
every two terms $t_1$ and $t_2$ such that $\isterm{\Ga}{t_1}{T_1}$,
  $\isterm{\Ga}{t_2}{T_2}$ and $t_1 \sim t_2$, there exists $p$ such
  that
  $\isterm{\Ga} {p} {\Heq{{T_1}} {{t_1}} {{T_2}} {{t_2}}}$.
%
However, during the proof, variables of different but (propositionally) equal
types are introduced and the context cannot be maintained to be the same
for both $t_1$ and $t_2$. Therefore, the translation needs to keep
track of this duplication of variables, plus a proof that they are
heterogeneously equal.
%
This mechanism is similar to what happens in the (relational) internal
parametricity translation in ITT introduced by
\cite{bernardy2012proofs} and recently rephrased in the setting of
\TemplateCoq~\cite{DBLP:conf/itp/AnandBCST18}. Namely, a context is not
translated as a telescope of variables, but as a telescope of triples
consisting of two variables plus a witness that they are in the
parametric relation.
%
In our setting, this amounts to consider telescope of triples
consisting of two variables plus a witness that they are
heterogeneously equal. We can express this by considering the
following dependent sums:
\[
\Pack{A_1}{A_2} := \Sum{x:A_1} \Sum{y:A_2} \Heq{A_1}{x}{A_2}{y}.
\]
%
This presentation inspired by the parametricity translation is crucial
in order to get an effective translation, because it is necessary to
keep track of the evolution of contexts when doing the translation on
open terms.
%
This ingredient is missing in Oury's work~\cite{oury2005extensionality},
which prevents him from deducing an effective (\ie constructive and
computable) translation from his theorem.

\section{Definitions of Extensional and Intensional Type Theories}
\label{sec:syntax-features}

This section presents the common syntax, typing and main properties of
ETT and ITT. Our type theories feature a universe hierarchy, dependent
products and sums as well as Martin Löf's identity types.

\subsection{Syntax of ETT and ITT}
\label{sec:ettsyntax}

The common syntax of ETT and ITT is given in Figure~\ref{fig:syntax}.
%
It features: dependent products $\Prod{x:A} B$, with (annotated)
$\lambda$-abstractions and (annotated) applications, negative dependent sums
$\Sum{x:A} B$ with (annotated) projections, sorts $\Ty{i}$, identity types
$\Eq{A}{u}{v}$ with reflection and elimination as well as terms
realising UIP and FunExt. Annotating terms with
otherwise computationally irrelevant typing information is a common
practice when studying the syntax of type theory precisely (see
\cite{streicher1993investigations} for a similar example).
%
We will write $A \to B$ for $\Prod{\_:A} B$ the non-dependent product /
function type.

We consider a fixed universe hierarchy without cumulativity, which
ensures in particular uniqueness of typing~\eqref{lem:uniq} which is
important for the translation.

\begin{figure*}
  \hrulefill
  \[
  \begin{array}{l@{~\,}r@{~\,}l@{\quad}l}
    s &\bnf& \Ty{i} ~(i \in \mathbb{N}) &\mbox{sorts (universes)} \\
    T,A,B,t,u,v &\bnf& x \bnfor \lam{x:A}{B} t \bnfor \app{t}{x:A}{B}{u}
    \bnfor \Prod{x:A} B \bnfor s
    &\mbox{dependent $\lambda$-calculus} \\
    &\bnfor& \pair{x:A}{B}{u}{v} \bnfor \pio{x:A}{B}{p} \bnfor \pit{x:A}{B}{p}
    \bnfor \Sum{x:A} B
    &\mbox{dependent pairs} \\
    &\bnfor& \refl{A} u \bnfor \J{A}{u}{x.e.P}{w}{v}{p} \bnfor \Eq{A}{u}{v}
    &\mbox{propositional equality} \\
    &\bnfor& \funext{x:A}{B}{f}{g}{e} \bnfor \uip{A}{u}{v}{p}{q}
    &\mbox{equality axioms} \\
    \Ga, \D &\bnf& \ctxempty \bnfor \Ga, x:A &\mbox{contexts}
  \end{array}
  \]
  \hrulefill
  \vspace{-2ex}
  \caption{Common syntax of ETT and ITT}
  \label{fig:syntax}
\end{figure*}

\paragraph{About Annotations.}
Although it may look like a technical detail, the use of annotation is more
fundamental in ETT than it is in ITT (where it is irrelevant and doesn't affect
the theory). And this is actually one of the main differences between our work
(and that of Martin \cite{hofmann1995conservativity} who has a similar
presentation) and the work of \cite{oury2005extensionality}.

Indeed, by using the standard model where types are interpreted as
cardinals rather than sets, it is possible to see that the equality
$\nat \to \nat = \nat \to \bool$ is independent from the theory, it is
thus possible to assume it (as an axiom, or for those that would still
not be convinced, simply under a $\lambda$ that would introduce this
equality).  In that context, the identity map $\lambda(x : \nat).\ x$
can be given the type $\nat \to \bool$ and we thus type
$(\lambda(x : \nat).\ x)\ \zero : \bool$.  Moreover, the
$\beta$-reduction of the non-annotated system used by Oury concludes
that this expression reduces to $\zero$, but cannot be given the type
$\bool$ (as we said, the equality $\nat \to \nat = \nat \to \bool$ is
independent from the theory, so the context is consistent). This means
we lack subject reduction in this case (or uniqueness of types,
depending on how we see the issue).  Our presentation has a blocked
$\beta$-reduction limited to matching annotations:
$\app{(\lam{x:A}{B}\ t)}{x:A}{B}{u} = t[x \sto u]$, from which subject
reduction and uniqueness of types follow.

Although subtle, this difference is responsible for Oury's need for an
extra axiom. Indeed, to treat the case of equality of applications in
his proof, he needs to assume the congruence rule for heterogeneous
equality of applications, which is not provable when formulated with
John Major equality (Fig.~\ref{fig:jmapp}). Thanks to annotations and
our notion of heterogeneous equality, we can prove this congruence
rule for applications.

\begin{figure}[h]
  \begin{mathpar}
    \infer[JMAPP]
    {\Heq{\forall (x : U_1). V_1}{f_1}{\forall (x : U_2). V_2}{f_2}\\
      \Heq{U_1}{u_1}{U_2}{u_2}}
    {\Heq{V_1[x \leftarrow u_1]}{f_1\ u_1}{V_2[x \leftarrow u_2]}{f_2\ u_2}}
    %
  \end{mathpar}
  \caption{Congruence of heterogeneous equality}
  \label{fig:jmapp}
\end{figure}

\subsection{The Typing Systems}

As usual in dependent type theory, we consider contexts which are
telescopes whose declarations may depend on any variable already
introduced. We note $\isterm{\Ga}{t}{A}$ to say that $t$ has type $A$
in context $\Ga$. $\istype{\Ga}{A}$ shall stand for
$\isterm{\Ga}{A}{s}$ for some sort $s$ and similarly
$\eqtype{\Ga}{A}{B}$ stands for $\eqtype{\Ga}{A}{B} : s$.

We use two relations $(s,s') \in \Ax$ (written $(s,s')$ for short)
and $(s,s',s'') \in \Rl$ (written $(s,s',s'')$) to constrain
the sorts in the typing rules for universes, dependent products and
dependent sums, as is done in any Pure Type System (PTS).
%
In our case, because we do not have cumulativity, the rules are as
follows:
%
\begin{mathpar}
  (\Ty{i}, \Ty{i+1}) \in \Ax

  (\Ty{i}, \Ty{j}, \Ty{\nmax{i}{j}}) \in \Rl
\end{mathpar}

\input{figty}

We give the typing rules of ITT in
Figure~\ref{fig:ty-rules}.
%
The rules are standard and we do not explain them.
%
Let us just point out the conversion rule, which says that $u:A$ can
be given the type $u:B$ when $A \equiv B$, \ie when $A$ and $B$ are
convertible.
%
As the notion of conversion is central in our work---the conversion of
ETT being translated to an equality in ITT---we provide an exhaustive
definition of it, with computational conversion rules (including
$\beta$-conversion or reduction of the elimination principle of
equality over reflexivity, see Figure~\ref{fig:conv-rules}), however
congruence conversion rules can be found in Appendix~\ref{sec:more-rules}
(Figure~\ref{fig:cong-rules}).
Note that we use Christine Paulin-Möhring's variant of the J rule rather than
Martin-Löf's original formulation.
%
Although pretty straightforward, being precise here is very important,
as for instance the congruence rule for $\lambda$-terms is the reason
why FunExt is derivable in ETT. Congruence of
equality terms is a standard extension of congruence to the new
principles we add (UIP and FunExt).


ETT is thus simply an extension of ITT (we write $\vdash_\exmark$ for
the associated typing judgment) with the reflection rule on equality,
which axiomatises that propositionally equal terms are convertible
(see Equation~\ref{eq:reflection}).
Note that, as already mentioned, in the presence of reflection and
$\mathsf{J}$, UIP is derivable so we could remove it from ETT, but
keeping it allows us to share a common syntax which makes the
statements of theorems simpler and does not affect the development.


\input{figconv}

\subsection{General Properties of ITT and ETT}

We now state the main properties of both ITT and ETT.
%
We do not detail their proof as they are standard and can be found
in the \Coq formalisation.

First, although not explicit in the typing system, weakening is
admissible in ETT and ITT.

\begin{lemma}[Weakening]
  \label{lem:weak}
  If $\isjg{\Ga}{J}$ and $\D$ extends $\Ga$ (possibly interleaving variables)
  then $\isjg{\D}{J}$.
\end{lemma}

Then, as mentioned above, the use of a non-cumulative hierarchy allows
us to prove that a term $t$ can be given at most one type in a
context $\Ga$, up-to conversion.

\begin{lemma}[Uniqueness of typing]
  \label{lem:uniq}
  If $\isterm{\Ga}{u}{T_1}$ and $\isterm{\Ga}{u}{T_2}$
  then $\eqtype{\Ga}{T_1}{T_2}$.
\end{lemma}

Finally, an important property of the typing system (seen as a mutual
inductive definition) is the possibility to deduce hypotheses from
their conclusion, thanks to inversion of typing. Note that it is
important here that our syntax is annotated for applications and
projections as it provides a richer inversion principle.

\begin{lemma}[Inversion of typing]
  \label{lem:inversion}
  \leavevmode
  \begin{enumerate}
    \item If $\isterm{\Ga}{x}{T}$ then $(x : A) \in \Ga$ and
    $\eqtype{\Ga}{A}{T}$.
    \item If $\isterm{\Ga}{\Ty{i}}{T}$ then $\eqtype{\Ga}{\Ty{i+1}}{T}$.
    \item If $\isterm{\Ga}{\Prod{x:A}{B}}{T}$ then $\isterm{\Ga}{A}{s}$ and
    $\isterm{\Ga, x:A}{B}{s'}$ and $\eqtype{\Ga}{s''}{T}$ for some $(s,s',s'')$.
    \item If $\isterm{\Ga}{\lam{x:A}{B} t}{T}$ then $\isterm{\Ga}{A}{s}$ and
    $\isterm{\Ga, x:A}{B}{s'}$ and $\isterm{\Ga, x:A}{t}{B}$ and
    $\eqtype{\Ga}{\Prod{x:A}{B}}{T}$.
    \item If $\isterm{\Ga}{\app{u}{x:A}{B}{v}}{T}$ then $\isterm{\Ga}{A}{s}$ and
    $\isterm{\Ga, x:A}{B}{s'}$ and $\isterm{\Ga}{u}{\Prod{x:A}{B}}$ and
    $\isterm{\Ga}{v}{A}$ and $\eqtype{\Ga}{B[ x \sto u]}{T}$.
    \item \ldots\ Analogous for the remaining term and type constructors.
  \end{enumerate}
\end{lemma}

\begin{proof}
  Each case is proven by induction on the derivation
  (which corresponds to any number of applications of the conversion rule
  following one introduction rule).
\end{proof}

\section{Relating Translated Expressions}
\label{sec:relation}

We want to define a relation on terms that equates two terms that are
the same up to transport.
%
This begs the question of what notion of transport is going to be
used.
%
Transport can be defined from elimination of equality as follows:
%
\begin{definition}[Transport]
  Given $\isterm{\Ga}{p}{\Eq{s}{T_1}{T_2}}$ and
  $\isterm{\Ga}{t}{T_1}$ we define the transport of $t$ along $p$, written
  $\translpo{p}\ t$, as
  $\app
    {\J
      {s}
      {T_1}
      {X.e.\ T_1 \to X}
      {\lam{x:T_1}{T_1} x}
      {T_2}
      {p}
    }
    {T_1}
    {T_2}
    {t}
  $ such that $\isterm{\Ga}{\translpo{p}\ t}{T_2}$.
\end{definition}
%
However, in order not to confuse the transports added by the
translation with the transports that were already present in the
source, we consider $\translpo{p}$ as part of the syntax in the
reasoning. It will be unfolded to its definition only after the
complete translation is performed.
%
This idea is not novel as Hofmann already had a $\mathsf{Subst}$ operator that
was part of his ITT (noted TT\textsubscript{I} in his
paper~\cite{hofmann1995conservativity}).

%
We first define the (purely syntactic) relation $\ir$ between ETT terms
and ITT terms in Figure~\ref{fig:ir-def} stating that the ITT term is
simply a decoration of the first term by transports. Its purpose is to
state how close to the original term its translation is. Then, we extend
this relation to a similarity relation $\sim$ on ETT terms by taking its
symmetric and transitive closure:
\begin{center}$\sim \coloneqq (\ir \cup \ir^{-1})^+$
\end{center}

\begin{lemma}[$\sim$ is an equivalence relation]
  \label{lem:sim-er}
  $\sim$ is reflexive, symmetric and transitive.
\end{lemma}

\begin{proof}
  For reflexivity we proceed by induction on the term.
\end{proof}

\input{figir}
%\input{figsim}

The goal is to prove that two terms in this relation, that are well-typed in the
target type theory, are heterogeneously equal. As for this notion, we recall
the definition we previously gave:
$\Heq{T}{t}{U}{u} := \Sum{p:\Eq{}{T}{U}} \Eq{}{\translpo{p}\ t}{u}$.
%
This definition of heterogeneous equality can be shown to be
reflexive, symmetric and transitive. Because of UIP, heterogeneous
equality collapses to equality when taken on the same type.

\begin{lemma}
  \label{lem:uip-cong}
  If $\isterm{\Ga}{e}{\Heq{A}{u}{A}{v}}$
  then there exists $p$ such that $\isterm{\Ga}{p}{\Eq{A}{u}{v}}$.
\end{lemma}

\begin{proof}
  This holds thanks to UIP on equality, which implies K, and so the
  proof of $A = A$ can be taken to be reflexivity.
\end{proof}

\begin{remark}
  As a corollary, $\Heqs$ on types corresponds to equality.
  Indeed when we have $\isterm{\Ga}{e}{\Heq{s}{A}{s'}{B}}$ we have
  that $\Eq{}{s}{s'}$, which implies that $s$ and $s'$ have the same sort
  and thus are syntactically the same (by an inversion argument).
\end{remark}

Before we can prove the fundamental lemma stating that two terms in relation
are heterogeneously equal, we need to consider another construction.
%
As explained in the introduction, when proving the property by
induction on terms, we introduce variables in the context that are
equal only up-to heterogeneous equality.
%
This phenomenon is similar to what happens in the parametricity
translation~\cite{bernardy2012proofs}.
%
Our fundamental lemma on the decoration relation $\sim$ assumes two
related terms of potentially different types $T1$ and $T2$ to produce an
heterogeneous equality between them. For induction to go through under
binders (e.g. for dependent products and abstractions), we hence need to
consider the two terms under different, but heterogeneously equal
contexts.
%
Therefore, the context we produce will not only be a telescope of
variables, but rather a telescope of triples consisting of two variables
of possibly different types, and a witness that they are heterogeneously
equal.
%
To make this precise, we define the following macro:
%
\[
\Pack{A_1}{A_2} := \Sum{x:A_1} \Sum{y:A_2} \Heq{}{x}{}{y}
\]
together with its projections
\begin{mathpar}
  \ProjO{p} := \pio{}{}{p}

  \ProjT{p} := \pio{}{}{\pit{}{}{p}}

  \ProjE{p} := \pit{}{}{\pit{}{}{p}}.
\end{mathpar}
%
We can then extend this notion canonically to contexts of the same
length that are well formed using the same sorts:
%
\[
\begin{array}{l}
    \Pack{(\Ga_1, x:A_1)}{(\Ga_2, x:A_2)} := \\
    (\Pack{\Ga_1}{\Ga_2}),
    x : \Pack{(\llift{\gamma}{}{A_1})}{(\rlift{\gamma}{}{A_2})} \\
    \\
    \Pack{\ctxempty}{\ctxempty} := \ctxempty.
\end{array}
\]
%
When we pack contexts, we also need to apply the correct projections for
the types in that context to still make sense. Assuming two contexts
$\Ga_1$ and $\Ga_2$ of the same length, we can define left and right
substitutions:
\[
\begin{array}{ll}
  \gamma_1 &:= [ x \leftarrow \ProjO{x}\ |\ (x : \_) \in \Ga_1 ] \\
  \gamma_2 &:= [ x \leftarrow \ProjT{x}\ |\ (x : \_) \in \Ga_2 ].
\end{array}
\]
These substitutions implement lifting of terms to packed contexts:
$\isterm{\Ga, \Pack{\Ga_1}{\Ga_2}}{\llift{\gamma}{}{t}}{\llift{\gamma}{}{A}}$
whenever $\isterm{\Ga, \Ga_1}{t}{A}$ (resp.
$\isterm{\Ga, \Pack{\Ga_1}{\Ga_2}}{\rlift{\gamma}{}{t}}{\rlift{\gamma}{}{A}}$
whenever $\isterm{\Ga, \Ga_2}{t}{A}$).

For readability, when $\Ga_1$ and $\Ga_2$ are understood we will write $\Gp$ for
$\Pack{\Ga_1}{\Ga_2}$.

Implicitly, whenever we use the notation $\Pack{\Ga_1}{\Ga_2}$ it means that
the two contexts are of the same length and well-formed with the same
sorts.
%
We can now state the fundamental lemma.

\begin{lemma}[Fundamental lemma]
  \label{lem:sim-cong}
  Let $t_1$ and $t_2$ be two terms. If $\isterm{\Ga, \Ga_1}{t_1}{T_1}$ and
  $\isterm{\Ga, \Ga_2}{t_2}{T_2}$ and $t_1 \sim t_2$ then there exists $p$ such
  that
  $\isterm{\Ga, \Pack{\Ga_1}{\Ga_2}}
          {p}
          {\Heq{\llift{\gamma}{}{T_1}}
               {\llift{\gamma}{}{t_1}}
               {\rlift{\gamma}{}{T_2}}
               {\rlift{\gamma}{}{t_2}}}$.
\end{lemma}

\begin{proof}
  The proof is by induction on the derivation of $t_1 \sim t_2$. We show
  the three most interesting cases:

  \begin{itemize}
  \item \textsc{Var}
    \[
      \infer[]
        { }
        {x \sim x}
      %
    \]
    If $x$ belongs to $\Ga$, we apply reflexivity---together with uniqueness of
    typing~\eqref{lem:uniq}---to conclude.
    Otherwise, $\ProjE{x}$ has the expected type (since
    $\llift{\gamma}{}{x} \equiv \ProjO{x}$ and $\rlift{\gamma}{}{x} \equiv \ProjT{x}$).

  \item \textsc{Application}
    \[
      \infer[]
        {t_1 \sim t_2 \\
         A_1 \sim A_2 \\
         B_1 \sim B_2 \\
         u_1 \sim u_2
        }
        {\app{t_1}{x:A_1}{B_1}{u_1} \sim \app{t_2}{x:A_2}{B_2}{u_2}}
      %
    \]
    We have $\isterm{\Ga, \Ga_1}{\app{t_1}{x:A_1}{B_1}{u_1}}{T_1}$ and
    $\isterm{\Ga, \Ga_2}{\app{t_2}{x:A_2}{B_2}{u_2}}{T_2}$ which means by
    inversion~\eqref{lem:inversion} that the subterms are well-typed.
    We apply the induction hypothesis and then conclude.
  \item \textsc{TransportLeft}
    \[
      \infer[]
        {t_1 \sim t_2}
        {\translpo{p}\ t_1 \sim t_2}
      %
    \]
    We have $\isterm{\Ga, \Ga_1}{\translpo{p}\ t_1}{T_1}$ and
    $\isterm{\Ga, \Ga_2}{t_2}{T_2}$.
    By inversion~\eqref{lem:inversion} we have
    $\isterm{\Ga, \Ga_1}{p}{\Eq{}{T_1'}{T_1}}$ and
    $\isterm{\Ga, \Ga_1}{t_1}{T_1'}$.
    By induction hypothesis we have $e$ such that
    $\isterm{\Ga, \Gp}{e}{\Heq{}{\llift{\gamma}{}{t_1}}{}{\rlift{\gamma}{}{t_2}}}$.
    From transitivity and symmetry we only need to provide a proof of
    $\Heq{}{\llift{\gamma}{}{t_1}}{}{\translpo{\llift{\gamma}{}{p}}\ \llift{\gamma}{}{t_1}}$ which is inhabited by
    $\pair{\_}{\_}{\llift{\gamma}{}{p}}{\refl{} (\translpo{\llift{\gamma}{}{p}}\ \llift{\gamma}{}{t_1})}$.
  \end{itemize}

  The complete proof can be found in Appendix~\ref{sec:proof-fund-lemma}.
\end{proof}

We can also prove that $\sim$ preserves substitution.

\begin{lemma}
  If $t_1 \sim t_2$ and $u_1 \sim u_2$ then
  $t_1[x \sto u_1] \sim t_2[x \sto u_2]$.
\end{lemma}

\begin{proof}
  We proceed by induction on the derivation of $t_1 \sim t_2$.
\end{proof}

\section{Translating ETT to ITT}
\label{sec:translation}

\subsection{The Translation}
\label{sec:the-translation}

We now define the translations (let us stress the plural here) of an
extensional judgment. We extend $\ir$ canonically to contexts
($\Ga \ir \Gb$ when they bind the same variables and the types are in
relation for $\ir$).

Before defining the translation, we define a set
$\transl{\xisterm{\Ga}{t}{A}}$ of typing judgments
in ITT associated to a typing judgment $\xisterm{\Ga}{t}{A}$ in ETT.
%
The idea is that this set describes all the possible translations that
lead to the expected property. When
$\isterm{\Gb}{\tb}{\Ab} \in \transl{\xisterm{\Ga}{t}{A}}$, we say that
$\isterm{\Gb}{\tb}{\Ab}$ realises $\xisterm{\Ga}{t}{A}$. The
translation will be given by showing that this set is inhabited by
induction on the derivation.

\begin{definition}[Characterisation of possible translations]
  \leavevmode
  \begin{itemize}
    \item For any $\xisctx{\Ga}$ we define $\transl{\xisctx{\Ga}}$ as a set of
    valid judgments (in ITT) such that
    $\isctx{\Gb} \in \transl{\xisctx{\Ga}}$ if and only if $\Ga \ir \Gb$.

    \item Similarly, $\isterm{\Gb}{\tb}{\Ab} \in \transl{\xisterm{\Ga}{t}{A}}$ iff
    $\isctx{\Gb} \in \transl{\xisctx{\Ga}}$ and $A \ir \Ab$ and $t \ir \tb$.
  \end{itemize}
\end{definition}

In order to better master the shape of the produced realiser, we state the
following lemma which shows that it has the same head
type constructor as the type it realises.
%
This is important for instance for the case of an application, where we
do not know a priori if the translated function has a dependent product
type, which is required to be able to use the typing rule for application.

\begin{lemma}
  \label{lem:choose}
  We can always \emph{choose} types $\Tb$ that have the same head constructor
  as $T$.
\end{lemma}

\begin{proof}
  Assume we have $\isterm{\Gb}{\tb}{\Tb} \in \transl{\xisterm{\Ga}{t}{T}}$.
  By definition of $\ir$,
  $T \ir \Tb$ means that $\Tb$ is shaped
  $\translpo{p}\ \translpo{q}\ ...\ \translpo{r}\ \Tb'$ with $\Tb'$ having
  the same head constructor as $T$. By inversion~\eqref{lem:inversion}, the
  subterms are typable, including $\Tb'$. Actually, from inversion, we
  even get that the type of $\Tb'$ is a universe. Then,
  using lemma~\ref{lem:sim-cong} and lemma~
  \ref{lem:uip-cong}, we get $\isterm{\Gb}{e}{\Eq{}{\Tb}{\Tb'}}$.
  We conclude with
  $\isterm{\Gb}{\translpo{e}\ \tb}{\Tb'} \in \transl{\xisterm{\Ga}{t}{T}}$.
\end{proof}

Finally, in order for the induction to go through, we need to know
that when we have a realiser of a derivation $\xisterm{\Ga}{t}{T}$, we can
pick an arbitrary other type realising $\xistype{\Ga}{T}$ and still
get a new derivation realising $\xisterm{\Ga}{t}{T}$ with that type.
%
This is important for instance for the case of an application, where
the type of the domain of the translated function may differ from the
type of the translated argument. So we need to be able to change it \textit{a
posteriori}.


\begin{lemma}
  \label{lem:change-type}
  When we have $\isterm{\Gb}{\tb}{\Tb} \in \transl{\xisterm{\Ga}{t}{T}}$
  and $\istype{\Gb}{\Tb'} \in \transl{\xistype{\Ga}{T}}$ then we also have
  $\isterm{\Gb}{\tb'}{\Tb'} \in \transl{\xisterm{\Ga}{t}{T}}$ for some $\tb'$.
\end{lemma}

\begin{proof}
  By definition we have $T \ir \Tb$ and $T \ir \Tb'$ and thus $T \sim \Tb$ and
  $T \sim \Tb'$, implying $\Tb \sim \Tb'$ by transitivity~\eqref{lem:sim-er}.
  By lemma~\ref{lem:sim-cong}
  (in the case $\Ga_1 \equiv \Ga_2 \equiv \ctxempty$) we get
  $\isterm{\Gb}{p}{\Heq{}{\Tb}{}{\Tb'}}$ for some $p$.
  By lemma~\ref{lem:uip-cong} (and lemma~\ref{lem:choose} to give
  universes as types to $\Tb$ and $\Tb'$) we can assume
  $\isterm{\Gb}{p}{\Eq{}{\Tb}{\Tb'}}$. Then
  $\isterm{\Gb}{\translpo{p}\ \tb}{\Tb'}$ is still a translation since $\ir$
  ignores transports.
\end{proof}

We can now define the translation. This is done by mutual induction on
context well-formedness, typing and conversion derivations. Indeed,
in order to be able to produce a realiser by induction, we need to show
that every conversion in ETT is translated as an heterogeneous equality
in ITT.

\begin{theorem}[Translation]
  \label{thm:translation}
  \leavevmode
  \begin{itemize}
    \item If\,\,\,$\xisctx{\Ga}$ then there exists
    $\isctx{\Gb} \in \transl{\xisctx{\Ga}}$,

    \item If\,\,\,$\xisterm{\Ga}{t}{T}$ then for any
    $\isctx{\Gb} \in \transl{\xisctx{\Ga}}$ there exist $\tb$ and $\Tb$ such
    that $\isterm{\Gb}{\tb}{\Tb} \in \transl{\xisterm{\Ga}{t}{T}}$,

    \item If\,\,\,$\xeqterm{\Ga}{u}{v}{A}$ then for any
    $\isctx{\Gb} \in \transl{\xisctx{\Ga}}$ there exist
    $A \ir \Ab, A \ir \Ab', u \ir \ub, v \ir \vb$ and $\eb$ such that
    $\isterm{\Gb}{\eb}{\Heq{\Ab}{\ub}{\Ab'}{\vb}}$.
  \end{itemize}
\end{theorem}

\begin{proof}
  We prove the theorem by induction on the derivation in the
  extensional type theory. We only show the two most interesting cases
  of application and conversion.
  The complete proof is given in Appendix~\ref{sec:corr-transl}.

  \begin{itemize}
    \item \textsc{Application}
    \[
      \infer[]
        {\xisterm{\Ga}{A}{s} \\
         \xisterm{\Ga,x:A}{B}{s'} \\
         \xisterm{\Ga}{t}{\Prod{x:A} B} \\
         \xisterm{\Ga}{u}{A}
        }
        {\xisterm{\Ga}{\app{t}{x:A}{B}{u}}{B[x \sto u]}}
      %
    \]
    Using IH together with lemmata~\ref{lem:choose} and~\ref{lem:change-type}
    we get $\isterm{\Gb}{\Ab}{s}$ and $\isterm{\Gb,x:\Ab}{\Bb}{s'}$ and
    $\isterm{\Gb}{\tb}{\Prod{x:\Ab} \Bb}$ and $\isterm{\Gb}{\ub}{\Ab}$
    meaning we can conclude
    $\isterm{\Gb}{\app{\tb}{x:\Ab}{\Bb}{\ub}}{\Bb[x \sto \ub]}
    \in \transl{\xisterm{\Ga}{\app{t}{x:A}{B}{u}}{B[x \sto u]}}$.

    \item \textsc{Conversion}
    \[
      \infer[]
        {\xisterm{\Ga}{u}{A} \\
         \xeqtype{\Ga}{A}{B}
        }
        {\xisterm{\Ga}{u}{B}}
      %
    \]
    By IH and lemma~\ref{lem:uip-cong} we have
    $\isterm{\Gb}{\eb}{\Eq{}{\Ab}{\Bb}}$ which implies
    $\istype{\Gb}{\Ab} \in \transl{\xistype{\Ga}{A}}$ by
    inversion~\eqref{lem:inversion}, thus, from lemma~\ref{lem:change-type}
    and IH we get $\isterm{\Gb}{\ub}{\Ab}$, yielding
    $\isterm{\Gb}{\translpo{\eb}\ \ub}{\Bb} \in \transl{\xisterm{\Ga}{u}{B}}$.
  \end{itemize}

\end{proof}


\subsection{Meta-theoretical Consequences}
\label{sec:meta-consequences}

We can check that all ETT theorems whose type are typable in ITT have
proofs in ITT as well:

\begin{corollary}[Preservation of ITT]
  \label{cor:preservation}
  If $\xisterm{}{t}{T}$ and $\istype{}{T}$ then there exist $\tb$ such that
    $\isterm{}{\tb}{T} \in \transl{\xisterm{}{t}{T}}$.
\end{corollary}
\begin{proof}
  Since $\isctx{\ctxempty} \in \transl{\xisctx{\ctxempty}}$, by
  Theorem~\eqref{thm:translation}, there exists $\tb$ and $\Tb$ such
  that
  $\isterm{}{\tb}{\Tb} \in \transl{\xisterm{}{t}{T}}$
  But as $\istype{}{T}$, we have
  $\istype{}{T} \in \transl{\xistype{}{T}}$, and,
  using Lemma~\ref{lem:change-type}, we obtain
  $\isterm{}{\tb}{T} \in \transl{\xisterm{}{t}{T}}$.
\end{proof}

\begin{corollary}[Relative consistency]
  \label{cor:consistency}
  Assuming ITT is consistent, there is no term $t$ such that
  $\xisterm{}{t}{\Prod{A:\Ty{0}}{A}}$.
\end{corollary}

\begin{proof}
  Assume such a $t$ exists. By the Corollary~\ref{cor:preservation},
  because $\istype{}{\Prod{A:\Ty{0}}{A}}$,
  there exists $\tb$ such that $\isterm{}{\tb}{\Prod{A:\Ty{0}}{A}}$ which
  contradicts the assumed consistency of ITT.
\end{proof}

\subsection{Optimisations}
\label{sec:optim}

Up until now, we remained silent about one thing: the size of the
translated terms. Indeed, the translated term is a decoration of the
initial one by transports which appear in many locations. For example,
at each application we use a transport by lemma \ref{lem:choose} to
ensure that the term in function position is given a function type. In
most cases---in particular when translating ITT terms---this produces
unnecessary transports (often by reflexivity) that we wish to avoid.

In order to limit the size explosion, in the above we use a different version of
transport, namely $\mathsf{transport}'$ such that
%
\begin{align*}
  \otransport{A_1}{A_2}{p}{t} &= t &\text{ when } A_1 =_\alpha A_2 \\
  &= \translpo{p}{t} &\text{ otherwise.}
\end{align*}
%
The idea is that we avoid \emph{trivially} unnecessary transports (we do not
deal with $\beta$-conversion for instance).
We extend this technique to the different constructors of equality (symmetry,
transitivity, \dots) so that they reduce to reflexivity whenever possible.
Take transitivity for instance:
%
\begin{align*}
  \otransitivity{\refl{} u}{q} &= q \\
  \otransitivity{p}{\refl{} u} &= p \\
  \otransitivity{p}{q} &= \translitivity{p}{q}.
\end{align*}
%
We show these \emph{defined terms} enjoy the same typing rules as their
counterparts and use them instead.
In practice it is enough to recover the exact same term when it is typed in ITT.

\section{Formalisation with Template-Coq}
\label{sec:form-with-templ}

We have formalised the translation
in the setting of \TemplateCoq~\cite{DBLP:conf/itp/AnandBCST18} in order to have
a more precise proof, but also to evidence the fact that the translation is
indeed constructive and can be used to perform computations.

\TemplateCoq is a \Coq library that has a representation of \Coq terms
as they are in \Coq's kernel (in particular using de Bruijn indices for
variables) and a (partial) implementation of the type checking algorithm
(not checking guardedness of fixpoints or positivity of inductive types).
%
It comes with a \Coq plugin that permits to quote \Coq terms into their
representations, and to produce \Coq terms from their representation
(if they indeed denote well-typed terms).
%
We have integrated our formalisation within that framework in order to
ensure our presentations of ETT and ITT are close to \Coq, but also to
take advantage of the quoting mechanism to produce terms using
the interactive mode (in particular we get to use tactics).
%
Note that we also rely on Mangin and Sozeau's
\Equations{}~\cite{DBLP:conf/itp/Sozeau10} plugin to derive nice
dependent induction principles.

Our formalisation takes full advantage of its easy interfacing with \TemplateCoq:
we define two theories, namely ETT and ITT, but ITT enjoys a lot of syntactic
sugar by having things such as transport, heterogeneous equality and packing as
part of the syntax. The operations regarding these constructors---in particular
the tedious ones---are written in \Coq and then quoted to finally be
\emph{realised} in the translation from ITT to \TemplateCoq.

\paragraph{Interoperability with \TemplateCoq.}
The translation we define from ITT to \TemplateCoq is not proven
correct, but it is not really important as it can just be seen as a
feature to observe the produced terms in a nicer setting. In any case,
\TemplateCoq does not yet provide a complete formalisation of CIC rules,
as guard checking of recursive definitions and strict positivity of
inductive type declarations are not formalised yet.

Our formalised theorems however do not depend on \TemplateCoq itself and as such
there is no need to \emph{trust} the plugin.

We also provide a translation from \TemplateCoq to ETT that we will describe
more extensively with the examples (Section~\ref{sec:examples}).

\subsection{Quick Overview of the Formalisation}

The file \rpath{SAst.v} contains the definition of the (common) abstract syntax
of ETT and ITT in the form of an inductive definition with de Bruijn
indices for variables (like in \TemplateCoq).
Sorts are defined separately in \rpath{Sorts.v} and we will address them later
in Section~\ref{sec:sorts}.

\begin{minted}{coq}
Inductive sterm : Type :=
| sRel (n : nat)
| sSort (s : sort)
| sProd (nx : name) (A B : sterm)
| sLambda (nx : name) (A B t : sterm)
| sApp (u : sterm) (nx : name) (A B v : sterm)
| sEq (A u v : sterm)
| sRefl (A u : sterm)
| (* ... *) .
\end{minted}

The files \rpath{ITyping.v} and \rpath{XTyping.v} define respectively the
typing judgments for ITT and ETT, using mutual inductive types.
Then, most of the files are focused on the meta-theory of ITT and can be ignored
by readers who don't need to see yet another proof of subject reduction.

The most interesting files are obviously those where the fundamental lemma
and the translation are formalised: \rpath{FundamentalLemma.v} and
\rpath{Translation.v}.
For instance, here is the main theorem, as stated in our formalisation:
%
\begin{minted}{coq}
Theorem complete_translation Σ :
  type_glob Σ ->
  (forall Γ (h : XTyping.wf Σ Γ), ∑ Γ', Σ |--i Γ' # ⟦ Γ ⟧ ) *
  (forall Γ t A (h : Σ ;;; Γ |-x t : A)
   Γ' (hΓ : Σ |--i Γ' # ⟦ Γ ⟧),
    ∑ A' t', Σ ;;;; Γ' |--- [t'] : A' # ⟦ Γ |--- [t] : A ⟧) *
  (forall Γ u v A (h : Σ ;;; Γ |-x u = v : A)
   Γ' (hΓ : Σ |--i Γ' # ⟦ Γ ⟧),
    ∑ A' A'' u' v' p', eqtrans Σ Γ A u v Γ' A' A'' u' v' p').
\end{minted}
%
Herein \mintinline{coq}{type_glob Σ} refers to the fact that some global
context is well-typed, its purpose is detailed in
Section~\ref{sec:inductives}.
The fact that the theorem holds in \Coq ensures we can actually
compute a translated term and type out of a derivation in ETT.

\subsection{Inductive Types and Recursion}
\label{sec:inductives}

In the proof of Section~\ref{sec:translation}, we didn't mention
anything about inductive types, pattern-matching or recursion as it is
a bit technical on paper.  In the formalisation, we offer a way to
still be able to use them, and we will even show how it works in
practice with the examples (Section\ref{sec:examples}).

The main guiding principle is that inductive types and induction are orthogonal
to the translation, they should more or less be translated to
themselves.
%
To realise that easily, we just treat an inductive definition as a way
to introduce new constants in the theory, one for the type, one for
each constructor, one for its elimination principle, and one equality
per computation rule.
%
For instance, the natural numbers can be represented by having the following
constants in the context:
%
\[
\begin{array}{l@{~}c@{~}l}
  \nat &:& \Ty{0} \\
  \zero &:& \nat \\
  \natsucc &:& \nat \to \nat \\
  \natrec &:& \forall P,\
  P\ \zero \to (\forall m,\ P\ m \to P\ (\natsucc\ m)) \to
  \forall n,\ P\ n \\
  \natrec_\zero &:& \forall P\ P_z\ P_s,\ \natrec\ P\ P_z\ P_s\ \zero = P_z \\
  \natrec_\natsucc &:& \forall P\ P_z\ P_s\ n,\\
  &&\natrec\ P\ P_z\ P_s\ (\natsucc\ n) = P_s\ n\ (\natrec\ P\ P_z\ P_s\ n)
\end{array}
\]
%
Here we rely on the reflection rule to obtain the computational behaviour of the
eliminator $\natrec$.

This means for instance that we do not consider inductive types that would only
make sense in ETT, but we deem this not to be a restriction and to the best of
our knowledge isn't something that is usually considered in the literature.
%
With that in mind, our translation features a global context of typed constants
with the restriction that the types of those constants should be well-formed
in ITT. Those constants are thus used as black boxes inside ETT.

With this we are able to recover what we were missing from
\Coq, without having to deal with the trouble of proving that the translation
doesn't break the guard condition of fixed points, and we are instead relying on
a more type-based approach.

\subsection{About Universes and Homotopy}
\label{sec:sorts}

The experienced reader might have noticed that our treatment of universes
(except perhaps for the absence of cumulativity) was really superficial and the
notion of sorts used is rather orthogonal to our main development.
This is even more apparent in the formalisation. Indeed, we didn't fix a
specific universe hierarchy, but instead specify what properties it should
have, in what is reminiscent to a (functional\sidenote{Meaning the sort of a
sort, and the sort of a product are functions, necessary to the uniqueness of
types~\eqref{lem:uniq}.}) PTS formulation.
%
\begin{minted}{coq}
Class Sorts.notion := {
  sort : Type ;
  succ : sort -> sort ;
  prod_sort : sort -> sort -> sort ;
  sum_sort : sort -> sort -> sort ;
  eq_sort : sort -> sort ;
  eq_dec : forall s z : sort, {s = z} + {s <> z} ;
  succ_inj : forall s z, succ s = succ z -> s = z
}.
\end{minted}
%
From the notion of sorts, we require functions to get the sort of a sort,
the sort of a product from the sorts of its arguments, and (crucially) the sort
of an identity type.
We also require some measure of decidable equality and injectivity on those.

This allows us to instantiate this by a lot of different notions including the
one presented earlier in the paper or even its extension with a universe $\Prop$
of propositions (like CIC~\cite{bertot2004interactive}). We present here two
instances that have their own interest.

\paragraph{$\Type$ in $\Type$.}
One of the instances we provide is one with only one universe $\Type$, with the
inconsistent typing rule $\Type : \Type$.
Although inconsistent, this allows us to interface with \TemplateCoq, without
the---for the time being---very time-consuming universe constraint checking.

\paragraph{Homotopy Type System and Two-Level Type Theory.}
Another interesting application (or rather instance) of our formalisation
is a translation from Homotopy Type System (HTS)~\cite{hts-sota} to
Two-Level Type Theory
(2TT)~\cite{DBLP:journals/corr/AltenkirchCK16,DBLP:journals/corr/AnnenkovCK17}.

HTS and 2TT arise from the incompatibility between UIP---recall it is provable
in ETT---and univalence. The idea is to have two distinct notions of equality
in the theory, a \emph{strict} one satisfying UIP, and a \emph{fibrant} one
corresponding to the homotopy type theory equality, possibly satisfying
univalence. This actually induces a separation in the types of the theory:
some of them are called \emph{fibrant} and the fibrant or homotopic equality
can only be eliminated on those.
HTS can be seen as an extension of 2TT with reflection on the strict equality
just like ETT is an extension of ITT.

We can recover HTS and 2TT in our setting by taking $\F{i}$ and $\Un{i}$ as
respectively the fibrant and strict universes of those theories
(for $i \in \mathbb{N}$), along with the following PTS rules:
%
\[
\begin{array}{l@{~}c@{~}l@{\qquad}l@{~}c@{~}l}
  (\F{i}, \F{i+1}) &\in& \Ax &
  (\Un{i}, \Un{i+1}) &\in& \Ax \\
  (\F{i}, \F{j}, \F{\nmax{i}{j}}) &\in& \Rl &
  (\F{i}, \Un{j}, \Un{\nmax{i}{j}}) &\in& \Rl \\
  (\Un{i}, \F{j}, \Un{\nmax{i}{j}}) &\in& \Rl &
  (\Un{i}, \Un{j}, \Un{\nmax{i}{j}}) &\in& \Rl \\
\end{array}
\]
%
and the fact that the sort of the (strict) identity type on $A : s$ is
the \emph{strictified} version of $s$, \ie $\Un{i}$ for $s = \Un{i}$ or
$s = \F{i}$.
In order to have the fibrant equality, one simply needs to do as in
Section~\ref{sec:inductives}.

In short, the translation from HTS to 2TT is basically the same as the one
from ETT to ITT we presented in this paper, and this fact is factorised through
our formalisation.

\subsection{ETT-flavoured \Coq: Examples}
\label{sec:examples}

In this section we demonstrate how our translation can bring extensionality to
the world of \Coq in action. The examples can be found in
\rpath{plugin\_demo.v}.

\paragraph{First, a pedestrian approach.}
%
We would like to begin by showing how one can write an example step by step
before we show how it can be instrumented and automated as a plugin.
For this we use a self-contained example without any inductive
types or recursion, illustrating a very simple case of reflection.
The term we want to translate is our introductory example of transport:
\[
  \lambda\ A\ B\ e\ x.\ x : \Pi\ A\ B.\ A = B \to
  A \to B
\]
which relies on the equality $e : A = B$ and reflection to convert $x
: A$ to $x : B$.
%
Of course, this definition isn't accepted in \Coq because this
conversion is not valid in ITT.
%
\begin{minted}{coq}
Fail Definition pseudoid (A B : Type) (e : A = B) (x : A) : B := x.
\end{minted}
%
However, we still want to be able to write it \emph{in some way}, in order to
avoid manipulating de Bruijn indices directly.
For this, we use a little trick by first defining a \Coq axiom to represent
an ill-typed term:
%
\begin{minted}{coq}
Axiom candidate : forall A B (t : A), B.
\end{minted}
%
\mintinline{coq}|candidate A B t| is a candidate \mintinline{coq}|t| of type
\mintinline{coq}|A| to inhabit type \mintinline{coq}|B|.
We complete this by adding a notation that is reminiscent to
\Agda's~\cite{norell2007towards} hole mechanism.
%
\begin{minted}{coq}
Notation "'{!' t '!}'" := (candidate _ _ t).
\end{minted}

We can now write the ETT function within \Coq.
%
\begin{minted}{coq}
Definition pseudoid (A B : Type) (e : A = B) (x : A) : B := {! x !}.
\end{minted}
%
We can then quote the term and its type to \TemplateCoq thanks to the
\mintinline{coq}|Quote Definition| command provided by the plugin.
%
\begin{minted}{coq}
Quote Definition pseudoid_term :=
  ltac:(let t := eval compute in pseudoid in exact t).
Quote Definition pseudoid_type :=
  ltac:(let T := type of pseudoid in exact T).
\end{minted}
%
The terms that we get are now \TemplateCoq terms, representing \Coq syntax.
We need to put them in ETT, meaning adding the annotations, and also removing
the \mintinline{coq}|candidate| axiom.
This is the purpose of the \mintinline{coq}|fullquote| function that we provide
in our formalisation.
%
\begin{minted}{coq}
Definition pretm_pseudoid :=
  Eval lazy in fullquote (2^18) Σ [] pseudoid_term empty empty nomap.
Definition tm_pseudoid :=
  Eval lazy in match pretm_pseudoid with
              | Success t => t
              | Error _ => sRel 0
              end.

Definition prety_pseudoid :=
  Eval lazy in fullquote (2^18) Σ [] pseudoid_type empty empty nomap.
Definition ty_pseudoid :=
  Eval lazy in match prety_pseudoid with
               | Success t => t
               | Error _ => sRel 0
               end.
\end{minted}
%
\mintinline{coq}|tm_pseudoid| and \mintinline{coq}|ty_pseudoid| correspond
respectively to the ETT representation of \mintinline{coq}|pseudoid| and its
type.
We then produce, using our home-brewed Ltac type-checking tactic, the
corresponding ETT typing derivation (notice the use of reflection to typecheck).
%
\begin{minted}{coq}
Lemma type_pseudoid : Σi ;;; [] |-x tm_pseudoid : ty_pseudoid.
Proof.
  unfold tm_pseudoid, ty_pseudoid.
  ettcheck. cbn.
  eapply reflection with (e := sRel 1).
  ettcheck.
Defined.
\end{minted}
%
We can then translate this derivation, obtain the translated term and then
convert it to \TemplateCoq.
%
\begin{minted}{coq}
Definition itt_pseudoid : sterm :=
  Eval lazy in
  let '(_ ; t ; _) :=
    type_translation type_pseudoid istrans_nil
  in t.

Definition tc_pseudoid : tsl_result term :=
  Eval lazy in
  tsl_rec (2 ^ 18) Σ [] itt_pseudoid empty.
\end{minted}
%
Once we have it, we \emph{unquote} the term to obtain a \Coq term
(notice that the only use of reflection has been replaced by a transport).
%
\begin{minted}{coq}
fun (A B : Type) (e : A = B) (x : A) => transport e x
     : forall A B : Type, A = B -> A -> B
\end{minted}

\paragraph{Making a Plugin with \TemplateCoq.}
%
All of this work is pretty systematic. Fortunately for us,
\TemplateCoq also features a monad to reify \Coq commands which we can
use to \emph{program} the translation steps.
As such we have written a complete procedure, relying on type checkers we
wrote for ITT and ETT, which can generate equality obligations.

Thanks to this, the user doesn't have to know about the details of
implementation of the translation, and stay within the \Coq ecosystem.

For instance, our previous example now becomes:
%
\begin{minted}{coq}
Definition pseudoid (A B : Type) (e : A = B) (x : A) : B := {! x !}.

Run TemplateProgram (Translate ε "pseudoid").
\end{minted}
%
This produces a \Coq term \mintinline{coq}{pseudoid'} corresponding to the
translation (\mintinline{coq}{ε} is the empty translation
context, see the next example to understand the need for a
translation context).
Notice how the user doesn't even have to provide any proof of equality or
derivations of any sort. The derivation part is handled by our own typechecker
while the obligation part is solved automatically by the \Coq obligation mechanism.

\paragraph{About inductive types.}
%
As we promised, our translation is able to handle inductive types.
For this consider the inductive type of vectors (or length-indexed lists) below,
together with a simple definition (we will remain in ITT for simplicity).
%
\begin{minted}{coq}
Inductive vec A : nat -> Type :=
| vnil : vec A 0
| vcons : A -> forall n, vec A n -> vec A (S n).

Arguments vnil {_}.
Arguments vcons {_} _ _ _.

Definition vv := vcons 1 _ vnil.
\end{minted}
%
This time, in order to apply the translation we need to extend the translation
context with \mintinline{coq}|nat| and \mintinline{coq}|vec|.
%
\begin{minted}{coq}
Run TemplateProgram (
    Θ <- TranslateConstant ε "nat" ;;
    Θ <- TranslateConstant Θ "vec" ;;
    Translate Θ "vv"
).
\end{minted}
%
The command \mintinline{coq}|TranslateConstant| enriches the current
translation context with the types of the inductive type and of its
constructors. The translation context then also contains associative
tables between our own representation of constants and those of \Coq.
Unsurprisingly, the translated \Coq term is the same as the original
term.

\paragraph{Reversal of vectors.}
%
Next, we tackle a motivating example: reversal on vectors.
Indeed, implementing this operation the same way it can
be done on lists ends up in the following conversion problem:
%
\begin{minted}{coq}
Fail Definition vrev {A n m} (v : vec A n) (acc : vec A m)
: vec A (n + m) :=
  vec_rect A (fun n _ => forall m, vec A m -> vec A (n + m))
           (fun m acc => acc)
           (fun a n _ rv m acc => rv _ (vcons a m acc))
           n v m acc.
\end{minted}
%
The recursive call returns a vector of length \mintinline{coq}|n + S m|
where the context expects one of length \mintinline{coq}|S n + m|. In ITT, these
types are not convertible. This example is thus a perfect fit for ETT where we
can use the fact that these two expressions always compute to the same thing
when instantiated with concrete numbers.
%
\begin{minted}{coq}
Definition vrev {A n m} (v : vec A n) (acc : vec A m)
: vec A (n + m) :=
  vec_rect A (fun n _ => forall m, vec A m -> vec A (n + m))
           (fun m acc => acc)
           (fun a n _ rv m acc => {! rv _ (vcons a m acc) !})
           n v m acc.

Run TemplateProgram (
    Θ <- TranslateConstant ε "nat" ;;
    Θ <- TranslateConstant Θ "vec" ;;
    Θ <- TranslateConstant Θ "Nat.add" ;;
    Θ <- TranslateConstant Θ "vec_rect" ;;
    Translate Θ "vrev"
).
\end{minted}
%
This generates four obligations that are all solved automatically. One of
them contains a proof of \mintinline{coq}|S n + m = n + S m| while the remaining
three correspond to the computation rules of addition (as mentioned before,
\mintinline{coq}|add| is simply a constant and does not compute in our
representation, hence the need for equalities).
%
The returned term is the following, with only one transport remaining
(remember our interpretation map removes unnecessary transports).
\begin{minted}{coq}
fun (A : Type) (n m : nat) (v : vec A n) (acc : vec A m) =>
vec_rect A
  (fun n _ => forall m, vec A m -> vec A (n + m))
  (fun m acc => acc)
  (fun a n₀ v₀ rv m₀ acc₀ =>
    transport (vrev_obligation_3 A n m v acc a n₀ v₀ rv m₀ acc₀)
      (rv (S m₀) (vcons a m₀ acc₀))) n v m acc
: forall A n m, vec A n -> vec A m -> vec A (n + m)
\end{minted}

\subsection{Towards an Interfacing between \Andromeda and \Coq}

\Andromeda~\cite{andromeda} is a proof assistant implementing ETT in a sense
that is really close to our formalisation. Aside from a concise nucleus with
a basic type theory, most things happen with the declaration of constants
with given types, including equalities to define the computational behaviour
of eliminators for instance.
This is essentially what we do in our formalisation.
Furthermore, their theory relies on $\Type : \Type$, meaning, our modular
handling of universes can accommodate for this as well.

All in all, it should be possible in the near future to use our translation
to produce \Coq terms out of \Andromeda developments.
%
Note that this would not suffer from the difficulties in generating typing
derivations since \Andromeda generates them.

\subsection{Composition with other Translations}

This translation also enables the formalisation of translations that
target ETT rather than ITT and still get mechanised proofs of (relative)
consistency by composition with this ETT to ITT translation.  This could
also be used to implement plugins based on the composition of
translations. In particular, supposing we have a theory which forms a
subset of ETT and whose conversion is decidable. Using this translation,
we could formalise it as an embedded domain-specific type theory and
provide an automatic translation of well-typed terms into witnesses in
\Coq. This would make it possible to extend conversion with the theory
of lists for example.

This would provide a simple way to justify the consistency of
CoqMT~\cite{DBLP:conf/lpar/JouannaudS17} for example, seeing it as an
extensional type theory where reflection is restricted to equalities
on a specific domain whose theory is decidable.


% \begin{comment}
\section{Extension: Translating HTS to 2TT}
\label{sec:extens-transl-hts}

\todo{Keep it as an easy extension? If so, we need to say more as this is no
longer a direct application.}

As we mentioned earlier, one of the interesting special cases covered by this
translation is the translation from 2-level type theory with reflection
(arguably a variant of Homotopy Type System~\cite{hts-sota} (HTS) to 2-level
type theory~\cite{altenkirch2016extending} (2TT).
%
We thus use two hierarchies $(\F{i})_{i \in \mathbb{N}}$, universes of fibrant
types, and $(\Un{i})_{i \in \mathbb{N}}$, universes of strict types.
We give the following axioms for the sorts:
%
\begin{align*}
  (\F{i}, \F{i+1}) &\in \Ax \\
  (\Un{i}, \Un{i+1}) &\in \Ax \\
  (\F{i}, \F{j}, \F{\nmax{i}{j}}) &\in \Rl \\
  (\F{i}, \Un{j}, \Un{\nmax{i}{j}}) &\in \Rl \\
  (\Un{i}, \F{j}, \Un{\nmax{i}{j}}) &\in \Rl \\
  (\Un{i}, \Un{j}, \Un{\nmax{i}{j}}) &\in \Rl
\end{align*}
%
We then equip our theory with a fibrant equality type defined as an inductive
type and the proof takes care of it. We can also add any inductive type that we
like, as for instance the natural numbers, the unit and empty types.
% \end{comment}

\section{Limitations and Axioms}
\label{sec:axioms}

Currently, the representation of terms and derivations and the
computational content of the proof only allow us to deal with the
translation of relatively small terms but we hope to improve that in
the future. As we have seen, the actual translation involves the
computational content of lemmata of inversion, substitution, weakening
and equational reasoning and thus cannot be presented as a simple
recursive definition on derivations.


As we already mentioned, the axioms K and FunExt are both
necessary in ITT if we want the translation to be conservative as they are
provable in ETT~\cite{hofmann1995conservativity}.
However, one might still be concerned about having axioms
as they can for instance hinder canonicity of the system.
In that respect, K isn't really a restriction since it preserves canonicity.
The best proof of that is probably \Agda itself which natively features K---in
fact, one needs to explicitly deactivate it with a flag if they wish to work
without.

The case of FunExt is trickier. It should be possible to realise
the axiom by composing our translation with a setoid
interpretation~\cite{altenkirch99} which validates it, or by going into a
system featuring it, for instance by implementing Observational Type
Theory~\cite{altenkirch2007observational} like
\Epigram~\cite{mcbride2004epigram}.

However, these two axioms are not used to define the translation itself,
but only to witness UIP and function extensionality in the translation to
\Coq.
The translation only relies on one axiom, called
\mintinline{coq}|conv_trans_AXIOM| in the formalisation, stating that conversion
of ITT is transitive.
%
The proof of this property basically sums up to the confluence of the
reduction rules of ITT which is out of scope for this paper and has
recently been formalised in Agda~\cite{Abel:2017:DCT:3177123.3158111} (in a
simpler setting with only one universe).
Regardless, this axiom inhabits a proposition (the type of conversion is in
\mintinline{coq}|Prop|) and is thus irrelevant for computation. Actually no
information about the derivation leaks to the production of the ITT term.

On a different note, the \mintinline{coq}|candidate| axiom allows us to derive
\mintinline{coq}|False| but is merely used to write ill-typed terms in \Coq.
The translated term will never make us of it and one can always check if a
term is relying on unsafe assumptions thanks to the
\mintinline{coq}|Print Assumptions| command.

\section{Related Works and Conclusion}
\label{sec:related-works}

The seminal works on the precise connection between ETT and ITT go
back to \cite{streicher1993investigations} and
\cite{hofmann1995conservativity,HofmannPhD}.
%
In particular, the work of Hofmann provides a categorical answer to
the question of consistency and conservativity of ETT over ITT with
UIP and FunExt.
%
Ten years later, \cite{oury2005extensionality,Oury2006} provided
a translation from ETT to ITT with
UIP and FunExt and other axioms (mainly due to
technical difficulties).
%
Although a first step towards a move from categorical semantics to a
syntactic translation, his work does not stress any constructive
aspect of the proof and shows that there merely exist translations in
ITT to a typed term in ETT.

\cite{van2013explicit} have later proposed and
formalised a similar translation between a PTS with and without explicit
conversion. This does not entail anything about ETT to ITT but we can
find similarities in that there is a witness of conversion between any
term and itself under an explicit conversion, which internalises
irrelevance of explicit conversions. This morally corresponds to a
Uniqueness of Conversions principle.

The Program \cite{sozeau:icfp07} extension of \Coq performs a
related coercion insertion algorithm, between objects in subsets on the
same carrier or in different instances of the same inductive family,
assuming a proof-irrelevance axiom. Inserting coercions locally is not
as general as the present translation from ETT to ITT which can insert
transports in any context.

In this paper we provide the first effective translation from ETT to ITT
with UIP and FunExt. The translation has been
formalised in \Coq using \TemplateCoq, a meta-programming plugin of
\Coq. This translation is also effective in the sense that we can
produce in the end a \Coq term using the \TemplateCoq denotation
machinery.
%
With ongoing work to extend the translation to the inductive fragment
of \Coq, we are paving the way to an extensional version of the \Coq
proof assistant which could be translated back to its intensional
version, allowing the user to navigate between the two modes, and in
the end produce a proof term checkable in the intensional fragment.

% \begin{acks}
  We would like to thank Andrej Bauer and Philipp Haselwarter with whom we had
  fruitful discussions on the subject, prior to this work.
  We also would like to thank the attendees of the Aarhus EUTypes 2018 meeting
  for their insightful feedback on the plugin stemming from the translation.
% \end{acks}

% \newpage
% \clearpage

\newpage
\onecolumn
\appendix

\section{Complementary rules}
\label{sec:more-rules}

\input{figcong}


\section{Proof of the fundamental lemma}
\label{sec:proof-fund-lemma}

\begin{lemma}[Fundamental lemma]
  Let $t_1$ and $t_2$ be two terms. If $\isterm{\Ga, \Ga_1}{t_1}{T_1}$ and
  $\isterm{\Ga, \Ga_2}{t_2}{T_2}$ and $t_1 \sim t_2$ then there exists $p$ such
  that
  $\isterm{\Ga, \Pack{\Ga_1}{\Ga_2}}
          {p}
          {\Heq{\llift{\gamma}{}{T_1}}
               {\llift{\gamma}{}{t_1}}
               {\rlift{\gamma}{}{T_2}}
               {\rlift{\gamma}{}{t_2}}}$.
\end{lemma}

For readability we will abbreviate the left and right substitutions
$\llift{\gamma}{}{\_}$ and $\rlift{\gamma}{}{\_}$ by $\upharpoonleft$
and $\upharpoonright$ respectively.

\begin{proof}
  We prove it by induction on the derivation of $t_1 \sim t_2$.

  \begin{itemize}
    \item \textsc{Var}
    \[
      \infer[]
        { }
        {x \sim x}
      %
    \]
    If $x$ belongs to $\Ga$, we apply reflexivity---together with uniqueness of
    typing~\eqref{lem:uniq}---to conclude.
    Otherwise, $\ProjE{x}$ has the expected type (since
    $\llift{\gamma}{}{x} \equiv \ProjO{x}$ and $\rlift{\gamma}{}{x} \equiv \ProjT{x}$).

    \item \textsc{TransportLeft}
    \[
      \infer[]
        {t_1 \sim t_2}
        {\translpo{p}\ t_1 \sim t_2}
      %
    \]
    We have $\isterm{\Ga, \Ga_1}{\translpo{p}\ t_1}{T_1}$ and
    $\isterm{\Ga, \Ga_2}{t_2}{T_2}$.
    By inversion~\eqref{lem:inversion} we have
    $\isterm{\Ga, \Ga_1}{p}{\Eq{}{T_1'}{T_1}}$ and
    $\isterm{\Ga, \Ga_1}{t_1}{T_1'}$.
    Then by induction hypothesis we have $e$ such that
    $\isterm{\Ga, \Gp}{e}{\Heq{}{\lo{t_1}}{}{\ro{t_2}}}$.
    From transitivity and symmetry we only need to provide a proof of
    $\Heq{}{\lo{t_1}}{}{\translpo{\lo{p}}\ \lo{t_1}}$ which is inhabited by
    $\pair{\_}{\_}{\lo{p}}{\refl{} (\translpo{\lo{p}}\ \lo{t_1})}$.

    \item \textsc{TransportRight}
    \[
      \infer[]
        {t_1 \sim t_2}
        {t_1 \sim \translpo{p}\ t_2}
      %
    \]
    Similarly.

    \item \textsc{Product}
    \[
      \infer[]
        {A_1 \sim A_2 \\
         B_1 \sim B_2
        }
        {\Prod{x:A_1} B_1 \sim \Prod{x:A_2} B_2}
      %
    \]
    We have $\isterm{\Ga, \Ga_1}{\Prod{x:A_1} B_1}{T_1}$ and
    $\isterm{\Ga, \Ga_2}{\Prod{x:A_2} B_2}{T_2}$ so by
    inversion~\eqref{lem:inversion} we have $\isterm{\Ga, \Ga_1}{A_1}{s_1}$ and
    $\isterm{\Ga, \Ga_1, x:A_1}{B_1}{s'_1}$ and
    $\eqtype{\Ga, \Ga_1}{s''_1}{T_1}$ for $(s_1,s'_1,s''_1) \in \Rl$
    (and similarly with $2$s).
    By induction hypothesis we have
    $\isterm{\Ga, \Gp}{p_A}{\Heq{}{\lo{A_1}}{}{\ro{A_2}}}$ and
    $\isterm
      {\Ga, \Gp, x : \Pack{A_1}{A_2}}
      {p_B}
      {\Heq{}{\lo{B_1}}{}{\ro{B_2}}}
    $
    hence the result (using UIP and FunExt, refer to the
    formalisation and especially to the file \rpath{Quotes.v} for more details
    on how to realise this equality).

    \item \textsc{Equality}
    \[
      \infer[]
        {A_1 \sim A_2 \\
         u_1 \sim u_2 \\
         v_1 \sim v_2
        }
        {\Eq{A_1}{u_1}{v_1} \sim \Eq{A_2}{u_2}{v_2}}
      %
    \]
    We have $\isterm{\Ga, \Ga_1}{\Eq{A_1}{u_1}{v_1}}{T_1}$ and
    $\isterm{\Ga, \Ga_2}{\Eq{A_2}{u_2}{v_2}}{T_2}$ so, by
    inversion~\eqref{lem:inversion}, we have
    $\isterm{\Ga, \Ga_1}{A_1}{s_1}$ and $\isterm{\Ga, \Ga_1}{u_1}{A_1}$ and
    $\isterm{\Ga, \Ga_1}{v_1}{A_1}$ as well as $\eqtype{\Ga, \Ga_1}{s_1}{T_1}$
    (and the same with $2$s). By induction hypothesis we thus have
    $\isterm{\Ga, \Gp}{p_{A}}{\Heq{}{A_1}{}{A_2}}$ and
    $\isterm{\Ga, \Gp}{p_u}{\Heq{}{u_1}{}{u_2}}$ and
    $\isterm{\Ga, \Gp}{p_v}{\Heq{}{v_1}{}{v_2}}$.
    We can thus conclude.

    \item \textsc{Reflexivity}
    \[
      \infer[]
        { }
        {s \sim s}
      %
    \]
    This one holds by reflexivity and uniqueness of typing~\eqref{lem:uniq}
    (indeed, $\lo{s} \equiv s$ and $\ro{s} \equiv s$).

    \item \textsc{Lambda}
    \[
      \infer[]
        {A_1 \sim A_2 \\
         B_1 \sim B_2 \\
         t_1 \sim t_2
        }
        {\lam{x:A_1}{B_1} t_1 \sim \lam{x:A_2}{B_2} t_2}
      %
    \]
    We have $\isterm{\Ga, \Ga_1}{\lam{x:A_1}{B_1} t_1}{T_1}$ and
    $\isterm{\Ga, \Ga_2}{\lam{x:A_2}{B_2} t_2}{T_2}$, thus, by
    inversion~\ref{lem:inversion} the subterms are well-typed and we can
    apply induction hypothesis. The conclusion follows similarly to the $\Pi$
    case.

    \item \textsc{Application}
    \[
      \infer[]
        {t_1 \sim t_2 \\
         A_1 \sim A_2 \\
         B_1 \sim B_2 \\
         u_1 \sim u_2
        }
        {\app{t_1}{x:A_1}{B_1}{u_1} \sim \app{t_2}{x:A_2}{B_2}{u_2}}
      %
    \]
    We have $\isterm{\Ga, \Ga_1}{\app{t_1}{x:A_1}{B_1}{u_1}}{T_1}$ and
    $\isterm{\Ga, \Ga_2}{\app{t_2}{x:A_2}{B_2}{u_2}}{T_2}$ which means by
    inversion~\eqref{lem:inversion} that the subterms are well-typed.
    We apply the induction hypothesis and then conclude.

    \item \textsc{Reflexivity}
    \[
      \infer[]
        {A_1 \sim A_2 \\
         u_1 \sim u_2
        }
        {\refl{A_1} u_1 \sim \refl{A_2} u_2}
      %
    \]
    We have $\isterm{\Ga, \Ga_1}{\refl{A_1} u_1}{T_1}$ and
    $\isterm{\Ga, \Ga_2}{\refl{A_2} u_2}{T_2}$ so by
    inversion~\eqref{lem:inversion} we have $\isterm{\Ga, \Ga_1}{A_1}{s_1}$ and
    $\isterm{\Ga, \Ga_1}{u_1}{A_1}$
    (same with $2$s). By IH we have $\Heq{}{\lo{A_1}}{}{\ro{A_2}}$ and
    $\Heq{\lo{A_1}}{\lo{u_1}}{\ro{A_2}}{\ro{u_2}}$.
    The proof follows easily.

    \item \textsc{Funext}
    \[
      \infer[]
        {A_1 \sim A_2 \\
         B_1 \sim B_2 \\
         f_1 \sim f_2 \\
         g_1 \sim g_2 \\
         e_1 \sim e_2
        }
        {\funext{x:A_1}{B_1}{f_1}{g_1}{e_1}
         \sim \funext{x:A_2}{B_2}{f_2}{g_2}{e_2}
        }
      %
    \]
    Similar.

    \item \textsc{UIP}
    \[
      \infer[]
        {A_1 \sim A_2 \\
         u_1 \sim u_2 \\
         v_1 \sim v_2 \\
         p_1 \sim p_2 \\
         q_1 \sim q_2
        }
        {\uip{A_1}{u_1}{v_1}{p_1}{q_1} \sim \uip{A_2}{u_2}{v_2}{p_2}{q_2}}
      %
    \]
    Similar.

    \item \textsc{J}
    \[
      \infer[]
        {A_1 \sim A_2 \\
         u_1 \sim u_2 \\
         P_1 \sim P_2 \\
         w_1 \sim w_2 \\
         v_1 \sim v_2 \\
         p_1 \sim p_2
        }
        {\J{A_1}{u_1}{x.e.P_1}{w_1}{v_1}{p_1} \sim
         \J{A_2}{u_2}{x.e.P_2}{w_2}{v_2}{p_2}
        }
      %
    \]
    Similar.
  \end{itemize}
\end{proof}

\section{Correctness of the translation}
\label{sec:corr-transl}
\begin{theorem}[Translation]
  \leavevmode
  \begin{itemize}
    \item If $\xisctx{\Ga}$ then there exists
    $\isctx{\Gb} \in \transl{\xisctx{\Ga}}$,

    \item If $\xisterm{\Ga}{t}{T}$ then for any
    $\isctx{\Gb} \in \transl{\xisctx{\Ga}}$ there exist $\tb$ and $\Tb$ such that
    $\isterm{\Gb}{\tb}{\Tb} \in \transl{\xisterm{\Ga}{t}{T}}$,

    \item If $\xeqterm{\Ga}{u}{v}{A}$ then for any
    $\isctx{\Gb} \in \transl{\xisctx{\Ga}}$ there exist
    $A \ir \Ab, A \ir \Ab', u \ir \ub, v \ir \vb$ and $\eb$ such that
    $\isterm{\Gb}{\eb}{\Heq{\Ab}{\ub}{\Ab'}{\vb}}$.
  \end{itemize}
\end{theorem}

\begin{proof}
  We prove the theorem by induction on the derivation in the extensional
  type theory. In most cases we need to assume some $\Gb$, translation of the
  context, we will implicitly refer to $\Gb$ in such cases as the one given as
  hypothesis.
  \leavevmode
  \begin{itemize}
    \item \textsc{Empty}
    \[
      \infer[]
        { }
        {\xisctx{\ctxempty}}
      %
    \]
    We have $\isctx{\ctxempty} \in \transl{\xisctx{\ctxempty}}$.

    \item \textsc{Extend}
    \[
      \infer[]
        {\xisctx{\Ga} \\
         \xistype{\Ga}{A}
        }
        {\xisctx{\Ga, x:A}}
      (x \notin \Ga)
    \]
    By IH we have $\isctx{\Gb} \in \transl{\xisctx{\Ga}}$ and, using $\Gb$
    as well as lemma~\ref{lem:choose},
    $\isterm{\Gb}{\Ab}{s} \in \transl{\xisterm{\Ga}{A}{s}}$.
    Thus $\isctx{\Gb, x:\Ab} \in \transl{\xisctx{\Ga, x:A}}$.

    \item \textsc{Sort}
    \[
      \infer[]
        {\xisctx{\Ga}}
        {\xisterm{\Ga}{s}{s'}}
      (s,s')
    \]
    We have $\isterm{\Gb}{s}{s'} \in \transl{\xisterm{\Ga}{s}{s'}}$.

    \item \textsc{Product}
    \[
      \infer[]
        {\xisterm{\Ga}{A}{s} \\
         \xisterm{\Ga,x:A}{B}{s'}
        }
        {\xisterm{\Ga}{\Prod{x:A} B}{s''}}
      (s,s',s'')
    \]
    By IH and lemma~\ref{lem:choose} we have $\isterm{\Gb}{\Ab}{s}$,
    meaning $\isctx{\Gb,x:\Ab} \in \transl{\xisctx{\Ga, x:A}}$,
    and then $\isterm{\Gb,x:\Ab}{\Bb}{s'}$.
    We thus conclude
    $\isterm{\Gb}{\Prod{x:\Ab} \Bb}{s''} \in
    \transl{\xisterm{\Ga}{\Prod{x:A} B}{s''}}$.

    \item \textsc{Sigma}
    \[
      \infer[]
        {\xisterm{\Ga}{A}{s} \\
         \xisterm{\Ga,x:A}{B}{s'}
        }
        {\xisterm{\Ga}{\Sum{x:A} B}{s''}}
      (s,s',s'')
    \]
    Similar.

    \item \textsc{Equality}
    \[
      \infer[]
        {\xisterm{\Ga}{A}{s} \\
         \xisterm{\Ga}{u}{A} \\
         \xisterm{\Ga}{v}{A}
        }
        {\xisterm{\Ga}{\Eq{A}{u}{v}}{s}}
      %
    \]
    By IH and lemma~\ref{lem:choose} we have $\isterm{\Gb}{\Ab}{s}$,
    and---using lemma~\ref{lem:change-type}---we also have
    $\isterm{\Gb}{\ub}{\Ab}$ and $\isterm{\Gb}{\vb}{\Ab}$.
    Then
    $\isterm{\Gb}{\Eq{\Ab}{\ub}{\vb}}{s} \in
    \transl{\xisterm{\Ga}{\Eq{A}{u}{v}}{s}}$.

    \item \textsc{Variable}
    \[
      \infer[]
        {\xisctx{\Ga} \\
         (x : A) \in \Ga
        }
        {\xisterm{\Ga}{x}{A}}
      %
    \]
    We have $\isctx{\Gb} \in \transl{\xisctx{\Ga}}$ (as we assumed, this is not
    an instance of the induction hypothesis) and $(x : A) \in \Ga$.
    By definition of $\Ga \ir \Gb$ we also have some $(x : \Ab) \in \Gb$
    with $A \ir \Ab$, thus
    $\isterm{\Gb}{x}{\Ab} \in \transl{\xisterm{\Ga}{x}{A}}$.

    \item \textsc{Conversion}
    \[
      \infer[]
        {\xisterm{\Ga}{u}{A} \\
         \xeqtype{\Ga}{A}{B}
        }
        {\xisterm{\Ga}{u}{B}}
      %
    \]
    By IH and lemma~\ref{lem:uip-cong} we have
    $\isterm{\Gb}{\eb}{\Eq{}{\Ab}{\Bb}}$ which implies
    $\istype{\Gb}{\Ab} \in \transl{\xistype{\Ga}{A}}$ by
    inversion~\eqref{lem:inversion}, thus, from lemma~\ref{lem:change-type}
    and IH we get $\isterm{\Gb}{\ub}{\Ab}$, yielding
    $\isterm{\Gb}{\translpo{\eb}\ \ub}{\Bb} \in \transl{\xisterm{\Ga}{u}{B}}$.

    \item \textsc{Lambda}
    \[
      \infer[]
        {\xisterm{\Ga}{A}{s} \\
         \xisterm{\Ga,x:A}{B}{s'} \\
         \xisterm{\Ga,x:A}{t}{B}
        }
        {\xisterm{\Ga}{\lam{x:A}{B} t}{\Prod{x:A} B}}
      %
    \]
    By IH and lemma~\ref{lem:choose} we have $\isterm{\Gb}{\Ab}{s}$ and thus
    $\isctx{\Gb, x:\Ab} \in \transl{\xisctx{\Ga,x:A}}$, meaning we can apply IH
    and lemma~\ref{lem:choose} to the second hypothesis to get
    $\isterm{\Gb,x:\Ab}{\Bb}{s'} \in \transl{\xisterm{\Ga,x:A}{B}{s'}}$ and then
    IH and lemma~\ref{lem:change-type} to get
    $\isterm{\Gb,x:\Ab}{\tb}{\Bb} \in \transl{\xisterm{\Ga,x:A}{t}{B}}$.
    All of this yields
    $\isterm{\Gb}{\lam{x:\Ab}{\Bb} \tb}{\Prod{x:\Ab} \Bb}
    \in \transl{\xisterm{\Ga}{\lam{x:A}{B} t}{\Prod{x:A} B}}$.

    \item \textsc{Application}
    \[
      \infer[]
        {\xisterm{\Ga}{A}{s} \\
         \xisterm{\Ga,x:A}{B}{s'} \\
         \xisterm{\Ga}{t}{\Prod{x:A} B} \\
         \xisterm{\Ga}{u}{A}
        }
        {\xisterm{\Ga}{\app{t}{x:A}{B}{u}}{B[x \sto u]}}
      %
    \]
    Using IH together with lemmata~\ref{lem:choose} and~\ref{lem:change-type}
    we get $\isterm{\Gb}{\Ab}{s}$ and $\isterm{\Gb,x:\Ab}{\Bb}{s'}$ and
    $\isterm{\Gb}{\tb}{\Prod{x:\Ab} \Bb}$ and $\isterm{\Gb}{\ub}{\Ab}$
    meaning we can conclude
    $\isterm{\Gb}{\app{\tb}{x:\Ab}{\Bb}{\ub}}{\Bb[x \sto \ub]}
    \in \transl{\xisterm{\Ga}{\app{t}{x:A}{B}{u}}{B[x \sto u]}}$.

    \item \textsc{Pair}
    \[
      \infer[]
        {\xisterm{\Ga}{u}{A} \\
         \xisterm{\Ga}{A}{s} \\
         \xisterm{\Ga,x:A}{B}{s'} \\
         \xisterm{\Ga}{v}{B[x \sto u]}
        }
        {\xisterm{\Ga}{\pair{x:A}{B}{u}{v}}{\Sum{x:A}{B}}}
      %
    \]
    Using IH with lemmata~\ref{lem:choose} and~\ref{lem:change-type} we
    translate all the hypotheses to conclude
    $\isterm{\Gb}{\pair{x:\Ab}{\Bb}{\ub}{\vb}}{\Sum{x:\Ab}{\Bb}}
    \in \transl{\xisterm{\Ga}{\pair{x:A}{B}{u}{v}}{\Sum{x:A}{B}}}$.

    \item \textsc{Proj$_1$}
    \[
      \infer[]
        {\xisterm{\Ga}{p}{\Sum{x:A}{B}}}
        {\xisterm{\Ga}{\pio{x:A}{B}{p}}{A}}
      %
    \]
    Similar.

    \item \textsc{Proj$_2$}
    \[
      \infer[]
        {\xisterm{\Ga}{p}{\Sum{x:A}{B}}}
        {\xisterm{\Ga}{\pit{x:A}{B}{p}}{B[x \sto \pio{x:A}{B}{p}]}}
      %
    \]
    Similar.

    \item \textsc{Reflexivity}
    \[
      \infer[]
        {\xisterm{\Ga}{A}{s} \\
         \xisterm{\Ga}{u}{A}
        }
        {\xisterm{\Ga}{\refl{A} u}{\Eq{A}{u}{u}}}
      %
    \]
    By IH we have $\isterm{\Gb}{\ub}{\Ab}$ and thus
    $\isterm{\Gb}{\refl{\Ab} \ub}{\Eq{\Ab}{\ub}{\ub}} \in
    \transl{\xisterm{\Ga}{\refl{A} u}{\Eq{A}{u}{u}}}$.

    \item \textsc{J}
    \[
      \infer[]
        {\xisterm{\Ga}{A}{s} \\
         \xisterm{\Ga}{u,v}{A} \\
         \xisterm{\Ga, x:A, e:\Eq{A}{u}{x}}{P}{s'} \\
         \xisterm{\Ga}{p}{\Eq{A}{u}{v}} \\
         \xisterm{\Ga}{w}{P[x \sto u, e \sto \refl{A} u]}
        }
        {\xisterm
          {\Ga}
          {\J{A}{u}{x.e.P}{w}{v}{p}}
          {P[x \sto v, e \sto p]}
        }
      %
    \]
    By IH and lemma~\ref{lem:choose} we have $\isterm{\Gb}{\Ab}{s}$.
    From this and IH and lemma~\ref{lem:change-type} we have
    $\isterm{\Gb}{\ub,\vb}{\Ab}$. We can thus deduce
    $\isctx{\Gb,x:\Ab,e:\Eq{\Ab}{\ub}{x}} \in \transl{\Ga, x:A, e:\Eq{A}{u}{x}}$
    which in turn gives us
    $\isterm{\Gb, x:\Ab,e:\Eq{\Ab}{\ub}{x}}{\Pb}{s'}$.
    Similarly we also get $\isterm{\Gb}{\pb}{\Eq{\Ab}{\ub}{\vb}}$ and
    $\isterm{\Gb}{\wb}{\Pb[x \sto \ub, e \sto \refl{\Ab} \ub]}$.
    All of this allows us to conclude
    $\isterm
      {\Gb}
      {\J{\Ab}{\ub}{x.e.\Pb}{\wb}{\vb}{\pb}}
      {\Pb[x \sto \vb, e \sto \pb]}
    \in
    \transl{
      \xisterm
        {\Ga}
        {\J{A}{u}{x.e.P}{w}{v}{p}}
        {P[x \sto v, e \sto p]}
    }$.

    \item \textsc{Funext}
    \[
      \infer[]
        {\isterm{\Ga}{f,g}{\Prod{x:A} B} \\
         \isterm
           {\Ga}
           {e}
           {\Prod{x:A} \Eq{B}{\app{f}{x:A}{B}{x}}{\app{g}{x:A}{B}{x}}}
        }
        {\isterm{\Ga}{\funext{x:A}{B}{f}{g}{e}}{\Eq{}{f}{g}}}
      %
    \]
    Similar.

    \item \textsc{UIP}
    \[
      \infer[]
        {\xisterm{\Ga}{e_1,e_2}{\Eq{A}{u}{v}}}
        {\xisterm{\Ga}{\uip{A}{u}{v}{e_1}{e_2}}{\Eq{}{e_1}{e_2}}}
      %
    \]
    Similar.

    \item \textsc{Beta}
    \[
      \infer[]
        {\xisterm{\Ga}{A}{s} \\
         \xisterm{\Ga,x:A}{B}{s'} \\
         \xisterm{\Ga,x:A}{t}{B} \\
         \xisterm{\Ga}{u}{A}
        }
        {\xeqterm
          {\Ga}
          {\app
            {(\lam{x:A}{B} t)}
            {x:A}
            {B}
            {u}
          }
          {t[x \sto u]}
          {B[x \sto u]}
        }
      %
    \]
    From IH and the lemmata, we even get the conversion, we conclude using
    reflexivity.

    \item \textsc{Proj$_1$-Red}
    \[
      \infer[]
        {\xisterm{\Ga}{A}{s} \\
         \xisterm{\Ga}{u}{A} \\
         \xisterm{\Ga,x:A}{B}{s'} \\
         \xisterm{\Ga}{v}{B[x \sto u]}
        }
        {\xeqterm
          {\Ga}
          {\pio{x:A}{B}{\pair{x:A}{B}{u}{v}}}
          {u}
          {A}
        }
      %
    \]
    Likewise.

    \item \textsc{Proj$_2$-Red}
    \[
      \infer[]
        {\xisterm{\Ga}{A}{s} \\
         \xisterm{\Ga}{u}{A} \\
         \xisterm{\Ga,x:A}{B}{s'} \\
         \xisterm{\Ga}{v}{B[x \sto u]}
        }
        {\xeqterm
          {\Ga}
          {\pit{x:A}{B}{\pair{x:A}{B}{u}{v}}}
          {v}
          {B[x \sto u]}
        }
      %
    \]
    Likewise.

    \item \textsc{J-Red}
    \[
      \infer[]
        {\xisterm{\Ga}{A}{\Un{i}} \\
         \xisterm{\Ga}{u}{A} \\
         \xisterm{\Ga, x:A, e:\Eq{A}{u}{x}}{P}{\Un{j}} \\
         \xisterm{\Ga}{w}{P[x \sto u, e \sto \refl{A} u]}
        }
        {\xeqterm
          {\Ga}
          {\J{A}{u}{x.e.P}{w}{u}{\refl{A} u}}
          {w}
          {P[x \sto u, e \sto \refl{A} u]}
        }
      %
    \]
    Likewise.

    \item \textsc{Conv-Refl}
    \[
      \infer[]
        {\xisterm{\Ga}{u}{A}}
        {\xeqterm{\Ga}{u}{u}{A}}
      %
    \]
    We conclude from IH and reflexivity of $\Heqs$.

    \item \textsc{Conv-Sym}
    \[
      \infer[]
        {\xeqterm{\Ga}{u}{v}{A}}
        {\xeqterm{\Ga}{v}{u}{A}}
      %
    \]
    We conclude from IH and symmetry of $\Heqs$.

    \item \textsc{Conv-Trans}
    \[
      \infer[]
        {\xeqterm{\Ga}{u}{v}{A} \\
         \xeqterm{\Ga}{v}{w}{A}
        }
        {\xeqterm{\Ga}{u}{w}{A}}
      %
    \]
    We conclude from IH and transitivity of $\Heqs$.

    % \item \textsc{}
    % \[
    %   \infer[]
    %     {\xisterm{\Ga}{A}{s} \\
    %      \xisterm{\Ga, x:A}{B}{s'} \\
    %      \xisterm{\Ga}{f}{\Prod{x:A} B}
    %     }
    %     {\xeqterm
    %       {\Ga}
    %       {\lam{x:A}{B} \app{f}{x:A}{B}{x}}
    %       {f}
    %       {\Prod{x:A} B}
    %     }
    %   %
    % \]
    % This works like the computation cases.

    \item \textsc{Conv-Conv}
    \[
      \infer[]
        {\xeqterm{\Ga}{t_1}{t_2}{T_1} \\
         \xeqtype{\Ga}{T_1}{T_2}
        }
        {\xeqterm{\Ga}{t_1}{t_2}{T_2}}
      %
    \]
    By IH (and lemma~\ref{lem:uip-cong}) we have
    $\isterm{\Gb}{\eb}{\Heq{\Tb_1}{\tb_1}{\Tb'_1}{\tb_2}}$ and
    $\isterm{\Gb}{p}{\Eq{}{\Tb''_1}{\Tb_2}}$. Also from
    lemmata~\ref{lem:sim-cong} and~\ref{lem:uip-cong} we have
    $\Eq{}{\Tb'_1}{\Tb''_1}$ and $\Eq{}{\Tb_1}{\Tb''_1}$, meaning we get
    $\Eq{}{\Tb'_1}{\Tb_2}$ and $\Eq{}{\Tb_1}{\Tb_2}$.
    This allows us to conclude by transporting along the aforementioned
    equalities.

    \item \textsc{Conv-Prod}
    \[
      \infer[]
        {\xeqterm{\Ga}{A_1}{A_2}{s} \\
         \xeqterm{\Ga,x:A_1}{B_1}{B_2}{s'}
        }
        {\xeqterm{\Ga}{\Prod{x:A_1} B_1}{\Prod{x:A_2} B_2}{s''}}
      (s,s',s'')
    \]
    We conclude exactly like we did in the proof of lemma~\ref{lem:sim-cong}.

    \item All congruences hold like in proof of
    lemma~\ref{lem:sim-cong}.

    \item \textsc{Conv-Eq}
    \[
      \infer[]
        {\xisterm{\Ga}{e}{\Eq{A}{u}{v}}}
        {\xeqterm{\Ga}{u}{v}{A}}
      %
    \]
    By IH and lemma~\ref{lem:choose} we have
    $\isterm{\Gb}{\eb}{\Eq{\Ab}{\ub}{\vb}}
    \in \transl{\xisterm{\Ga}{e}{\Eq{A}{u}{v}}}$ which yields the conclusion we
    wanted.
  \end{itemize}
\end{proof}