% \setchapterpreamble[u]{\margintoc}
\chapter{Acknowledgements}
\labch{ack}

(There is no real order in which I'm citing people!)

You are about to read my thesis which I wrote during lockdown, an experience
that I survived together with Laura whom I'd like to thank deeply.

I would of course like to thank my family who have been a great support in my
life: my mom, Lancelot, Tess, Martin, Elena and my grandmother who left us before
I could finish this work.


This work has had many readers helping me find bugs and typos hidden---or not
so hidden---inside. I'd like to thank Meven, Loïc, Kenji, Xavier, Gaëtan and
Matthieu for their useful remarks, and particularly Pierre who helped me greatly
on writing my introductory chapters and Andreas who gave me many pointers on how
to improve this document.
Regarding the manuscript, I would like to thank the reviewers Andrej and Herman
for accepting to spend their time reading this long material and for the useful
comments they gave me.
I would also of course like to thank the whole jury with whom I had nice
interactions in the past and hopefully in the future: Andreas, Assia, Bas,
Gilles and Hugo.


I am very grateful to Nicolas and Matthieu for advising me for those three years
and even before when I was their intern. During my stay in Nantes I had very
fruitful and interesting interactions with everyone at the Gallinette team (past
and present, even briefly, even never): Xavier, Meven, Loïc, Joachim, Gaëtan,
Simon, Marie, Kenji, Julien, Assia, Guillaume, Hervé, Rémi, Nicolas, Matthieu,
Pierre-Marie, Guilhem, Ambroise, Étienne, Pierre, Maxime, Igor, Kevin, Benedikt,
Danil, Antoine, Enzo, Lucas and Cyril.
A special mention for Ambroise whom I thank for his thanking me in his own
thesis with the goal of having me thank him in return.


I would like to thank the teachers that led me to where I am today, namely
M. Marizy, M. Lyx, M. Gonzales, M. Marino, Ms. Chireux, and M. Alessandri.
I would also like to thank Paul who coached me during my preparatory class
and has been a very good friend of mine ever since.
I would like to thank everyone I met at ENS Cachan, in particular my teachers
(including teachers at the MPRI) that were a great influence on me. I will
mention my friends Mathias, Léo (and the other Léos), Théis, Félicien, Itsaka,
Pierre-Léo, Mathieu, Victor, Olivier, Julien, Pierre, Thomas, Charlie, Damien,
Paul and François.


I would like to thank Andreas and everyone I met at Chalmers who made me
discover dependent types and Sweden: Andrea, Anders, Inari, Jesper, Vincent,
Hans, Simon, Bassel, Guilhem, Peter, Ana, Thierry, Nils Anders, Léo, Victor
and Paul.
%
Thanks to everyone I met at Paris 7: Pierre, Théo, Cyril, Étienne, Maxime,
Jovana, Zeinab, Kostia.
%
I would like to thank Andrej and everyone I met at Ljubljana
(or from Ljubljana): Philipp, Niels, Brett, Matija, Alex, Anja and Peter.


I must have forgotten people, so I thank them, they will know.


...and you! (Please open a pull-request.)


% % From CPP
% I would like to thank Andrej Bauer and Philipp Haselwarter with whom I had
% fruitful discussions on the subject, prior to this work.
% I also would like to thank the attendees of the Aarhus EUTypes 2018 meeting
% for their insightful feedback on the plugin stemming from the translation.