% \setchapterpreamble[u]{\margintoc}
\chapter{Translations}
\labch{translations}

Syntactical translations are a special case of program transformations suited
for type theory, and a method of choice to get models of type theories.
This is better studied in Simon Boulier's
thesis~\sidecite{boulier17:next-syntac-model-type-theor,boulier2018extending}
but in order to keep this document self contained, I will do my best to give a
meaningful excerpt here.

\section{Syntactical translations}

\subsection{Definition}
\labsubsec{syn-trans-def}

A \emph{syntactical} translation, as the name suggests, operates on the syntax
of terms. It has a source theory \cS and a target theory \cT with possibly
different syntax and typing rules---though usually \cS is an extension of \cT,
more on that later.

A translation is then given as two functions:
\begin{itemize}
  \item \([.]\) taking \cS terms to \cT terms;
  \item \(\transl{.}\) taking \cS types to \cT types.
\end{itemize}
And \(\transl{.}\) is---often---canonically extended to contexts:
\[
\begin{array}{lcl}
  \transl{\ctxempty} &\coloneqq& \ctxempty \\
  \transl{\Ga, x : A} &\coloneqq& \transl{\Ga}, x : \transl{A}
\end{array}
\]
\marginnote[-2cm]{
  It is not always the case that the \(\transl{\ctxempty}\) is also the empty
  context, and sometimes one variable is translated to several, but in a lot of
  cases it can still be encoded that way.
}

Usually, \(\transl{.}\) is built using \([.]\) and sometimes corresponds
exactly to \([.]\).
The translation is often done in an homomorphic way.

There are two main properties expected of a translation: type preservation
and preservation of falsehood.

\begin{definition}[Type preservation]
  A translation \([.], \transl{.}\) is type preserving when
  \(\isterm{\Ga}{t}{A}\) implies \(\isterm{\transl{\Ga}}{[t]}{\transl{A}}\).
\end{definition}

\begin{definition}[Preservation of falsehood]
  A translation \([.], \transl{.}\) from \cS to \cT is said to preserve
  falsehood when the translation of the empty type implies falsehood in the
  target:
  \[ \vdash_\cT \transl{\bot_\cS} \to \bot_\cT \]
\end{definition}

If a translation satisfies both those properties, we can conclude consistency
of \cS relatively to \cT.

\begin{theorem}[Relative consistency]
  If there is a translation from \cS to \cT that preserves typing and
  falsehood, then the consistency of \cT implies that of \cS.
\end{theorem}

\begin{proof}
  Assume a translation \([.], \transl{.}\) from \cS to \cT that preserves typing
  and falsehood. We will prove the contraposition and assume inconsistency in
  \cS, that is the existence of a term \(t\) such that
  \[ \vdash_\cS t : \bot_\cS \]
  By preservation of typing, we get
  \[ \vdash_\cT [t] : \transl{\bot_\cS} \]
  Moreover, since the translation preserves falsehood, there is a term \(f\)
  such that \(\vdash_\cT f : \transl{\bot_\cS} \to \bot_\cT\).
  Thus, we conclude, using application that
  \[ \vdash_\cT f\ t : \bot_\cT \]
  in other words that \cT is also inconsistent.
\end{proof}

Relative consistency is all the more interesting when the source theory is more
complex than the target and even more so when \cT is a theory that we trust.

Type preservation can be achieved by proving intermediary properties:
substitutivity, preservation of reduction and of conversion.

\marginnote[0.6cm]{
  Here I write \(t\{ x \sto u\}\) for \(t\) where \(x\) is substituted by \(u\),
  rather than \(t[x \sto u]\) to avoid confusion with the translation operator
  also written with square brackets.
}
\begin{definition}[Substitutivity]
  A translation \([.], \transl{.}\) is substitutive when
  \[
    \begin{array}{lcl}
      [t\{x \sto u\}] &\coloneqq& [t]\{x \sto [u]\} \\
      \transl{A\{x \sto u\}} &\coloneqq& \transl{A}\{x \sto [u]\} \\
    \end{array}
  \]
\end{definition}

\begin{definition}[Preservation of reduction]
  A translation \([.], \transl{.}\) is said to preserve reduction when
  \(u \red v\) implies \([u] \red [v]\) (and \(\transl{u} \red \transl{v}\)).
\end{definition}

\begin{definition}[Preservation of conversion]
  A translation \([.], \transl{.}\) is said to preserve conversion when
  \(u \equiv v\) implies \([u] \equiv [v]\)
  (and \(\transl{u} \equiv \transl{v}\)).
\end{definition}

\subsection{Example: the \(\times\ \mathsf{bool}\) translation}

A simple yet non-trivial example of translation (or class of translations) is
the so-called \(\times\ \mathsf{bool}\) translation, again
from~\sidecite{boulier17:next-syntac-model-type-theor,boulier2018extending}.

As I said earlier, the source is going to be an extension of the target.
I will keep the target pretty basic, as a simple instance of \acrshort{ITT}
with a hierarchy of universes \(\Ty{i}\) (for \(i \in \mathbb{N}\)) with the
following rules:
\begin{itemize}
  \item \((\Ty{i}, \Ty{j}) \in \Ax\) for \(i \le j\);
  \item \((\Ty{i}, \Ty{j}, \Ty{k}) \in \Rl\) for \(i, j \le k\).
\end{itemize}
\marginnote[-1.6cm]{
  Refer to \frefsubsec{pts-itt} to see what theory we extend here.
}
Otherwise it has to feature a boolean type \(\bool\)---with the usual \(\ttrue\)
and \(\ffalse\)---product types \(A \times B\) as special
cases\sidenote{See \frefsec{inductive-types}.} of \(\Sigma\)-types,
and equality.
%
\begin{mathpar}
  \infer
    { }
    {\isterm{\Ga}{\bool}{\Ty{0}}}
  %

  \infer
    { }
    {\isterm{\Ga}{\ttrue}{\bool}}
  %

  \infer
    { }
    {\isterm{\Ga}{\ffalse}{\bool}}
  %
\end{mathpar}

The source is going to be the target extended with a new principle corresponding
to the negation of \acrlong{funext}, that is the existence of two functions that
are extensionally equal, but not equal themselves.
%
\[
  \infer
    { }
    {
      \isterm
        {\Ga}
        {\notfunext}
        {\Sigma\ A\ B\ (f\ g : A \to B).\
          (\forall x.\ f\ x = g\ x) \times f \not= g
        }
    }
  %
\]
\marginnote[-1.4cm]{
  Inequality \(x \not= y\) is defined as the negation of equality, that is a map
  to the empty type: \(x = y \to \bot\).
}
%
I will write \(\Notfunext\) for the type of \(\notfunext\) in the following.

The idea is to inhabit this extra principle in the target which does not feature
it, or rather to inhabit its translation \(\transl{\Notfunext}\).
In order to distinguish functions of the target, while keeping them
observationally equal, we translate them by adding a boolean label to them.
Basically, \(\lambda x. x\) is sent to \((\lambda x. x, \ttrue)\), the
\(\ttrue\) telling us the it might have come from the translation of a function
from \cS.
Any function coming with \(\ffalse\) instead will not be a translated term.

The translation is defined as follows:
%
\[
\begin{array}{lcl}
  [ x ] &\coloneqq& x \\
  \rowcolor{highlightcol} \relax
  [ \lambda (x : A).\ t ] &\coloneqq& (\lambda (x : \transl{A}).\ [t],\ttrue) \\
  \rowcolor{highlightcol} \relax
  [ t\ u ] &\coloneqq& [t].1\ [u] \\ \relax
  [ (u, v) ] &\coloneqq& ([u], [v]) \\ \relax
  [ p.1 ] &\coloneqq& [p].1 \\ \relax
  [ p.2 ] &\coloneqq& [p].2 \\ \relax
  [ \ttrue ] &\coloneqq& \ttrue \\ \relax
  [ \ffalse ] &\coloneqq& \ffalse \\ \relax
  [ \tif{t}{b. P}{u}{v} ] &\coloneqq& \tif{[t]}{b. \transl{P}}{[u]}{[v]} \\
  \relax
  [ \refl{A} u ] &\coloneqq& \refl{\transl{A}} [u] \\ \relax
  [ \J{A}{u}{x.e.P}{w}{v}{p} ] &\coloneqq&
  \J{\transl{A}}{[u]}{x.e.\transl{P}}{[w]}{[v]}{[p]} \\ \relax
  [ \Ty{i} ] &\coloneqq& \Ty{i} \\
  \rowcolor{highlightcol} \relax
  [ \Pi (x : A).\ B ] &\coloneqq&
  (\Pi (x : \transl{A}).\ \transl{B}) \times \bool \\ \relax
  [ \Sigma (x : A).\ B ] &\coloneqq& \Sigma (x:\transl{A}).\ \transl{B} \\
  \relax
  [ \bool ] &\coloneqq& \bool \\ \relax
  [ u =_A v ] &\coloneqq& [u] =_{\transl{A}} [v] \\ \relax
  \\
  \transl{A} &\coloneqq& [A]
\end{array}
\]
\marginnote[-0.6cm]{Notice how \(\transl{.}\) is just defined as \([.]\).}

I actually did not provide the full translation here since I \emph{forgot}
the definition of \([\notfunext]\).
Writing down the term is rather tedious and boring, so instead I will give an
argument as to why it exists: you can provide any type for \(A\) and \(B\), for
instance \(\Ty{0}\) and then for \(f\) and \(g\) we give
\((\lambda (x:\Ty{0}).\ x, \ttrue)\) and
\((\lambda (x:\Ty{0}).\ x, \ffalse)\) that will be pointwise equal (once
projected) but different nonetheless.

This translation satisfies all the properties I mentioned in
\arefsubsec{syn-trans-def}, including preservation of typing and of falsehood.
As such it shows that the negation of \acrshort{funext} is consistent with
\acrshort{ITT}. On the other hand, it can be shown that \acrshort{funext} is
also consistent, and as such independent from \acrshort{ITT}.
\reminder[-2.7cm]{Independence}{
  A proposition \(P\) is \emph{independent} from a theory \cT when both
  \(\cT + P\) and \(\cT + \neg P\) are consistent relatively to \cT.
}
For details, you should again refer to
\sidecite{boulier17:next-syntac-model-type-theor,boulier2018extending}
which gives a more comprehensive treatment of this and other translations.

\section{Other translations}

As I mentioned earlier, it is not alaways the case that contexts are translated
homomorphically. For instance, in the binary parametricity
translation~\sidecite{bernardy2012proofs} a variable
\(a : A\) in the context is sent to three variables:
\[
  a_0 : \transl{A}_0, a_1 : \transl{A}_1, a_\varepsilon : \transl{A}\ a_0\ a_1
\]
but that can still be encoded as a unique variable
\[
  a : \Sigma\ (x : \transl{A}_0)\ (y : \transl{A}_1).\ \transl{A}\ x\ y
\]

Sometimes, the empty context itself is not translated to the empty context
as is the case in the forcing translation~\sidecite{jaber2016definitional} and
such an encoding becomes impossible.
Even worse, in this document, in \arefpart{elim-reflection}, I am going to
present a translation that is not even syntactic! Indeed, sometimes the
translation can only be conducted at the derivation level: derivations in \cS
are sent to derivations in \cT. It often reflects the fact that terms in \cS are
not enough to recover the derivation when it exists\sidenote{In other words,
type-checking is undecidable.}.
We could however imagine performing the translation on some intermediary between
the judgment and its derivation, containing just the right amount of
information. This brings us back to the discussion on the notion of
proof.

In those cases, the definitions and properties have to be adapted. This usually
has to be done on a case-by-case basis as for now there is no general framework
to deal with all those kinds of translations.

\section{Conversative extensions}

Translations are a nice way to obtain models of type theories, in particular
when showing that a certain principle extending a theory can be given an
interpretation and even computational content.
\marginnote[-1cm]{
  I did not dwell on this but as the new principle is given a definition in the
  target, we can \emph{pull out} its computational behaviour from there while
  respecting preservation of reduction.
}

We can even be more restrictive on translations if we want to show that the new
principles are not adding new \emph{truths}, essentially saying that the
extension is more convenient while not really changing the logic.
There is an appropriate notion that is not specific to translations or even
type theory which is that of \emph{conservative extension}.

\begin{definition}[Conversative extension]
  An extension \(\cT_2\) of a theory \(\cT_1\) is \emph{conservative} if every
  theorem of \(\cT_2\) that can be stated in \(\cT_1\) is already a theorem of
  \(\cT_1\).
\end{definition}
\marginnote[-1.5cm]{
  An extension of a theory is a theory that proves at least as many theorems as
  the one it extends.
}

In the context of type theories it can be reformulated as follows.

\begin{definition}[Conservative extension of a type theory]
  An extension \(\cT_2\) of a type theory \(\cT_1\) is \emph{conservative} if
  for every judgement \(\vdash_2 t : A\) such that \(\vdash_1 A\)
  (\ie \(A\) is a type in \(\cT_1\)), there exists \(t'\) such that
  \(\vdash_1 t' : A\).
\end{definition}

We will call a translation \emph{conservative} when it allows to exhibit that
the source theory \cS is a conservative extension of the target \cT, typically
when every type of \cT (\ie \(\vdash_\cT A\))---which makes sense in \cS since
it extends \cT---is sent to itself (\(\transl{A} = A\)) or at least implies
itself:
\[
  \vdash_\cT \transl{A} \to A
\]
This is a kind of generalisation of preservation of falsehood.

The \(\times\ \bool\) translation is not conservative because
\[
  \vdash \transl{\Notfunext} \not\to \Notfunext
\]
and more generally \acrshort{ITT} extended with \(\Notfunext\) is not
conservative over \acrshort{ITT}.
I will present a conservative translation in \arefpart{elim-reflection}.

A simpler example could be that of encodings: say you want to add the booleans
\(\bool\) to a type theory that already features a unit type \(\unit\) and
disjunctions \(A + B\), booleans can be encoded as \(\unit + \unit\) so they
are not strictly necessary but having them as first-class objects can still be
convenient. A translation corresponding to the encoding is conservative and as
such we can keep working on the meta-theory of a minimal theory (without
the booleans).
%
\[
\begin{array}{lcl}
  [ x ] &\coloneqq& x \\ \relax
  [ \lambda (x : A).\ t ] &\coloneqq& \lambda (x : \transl{A}).\ [t] \\
  \relax
  [ t\ u ] &\coloneqq& [t]\ [u] \\
  \rowcolor{highlightcol} \relax
  [ \ttrue ] &\coloneqq& \inl\ \tunit \\
  \rowcolor{highlightcol} \relax
  [ \ffalse ] &\coloneqq& \inr\ \tunit \\
  \rowcolor{highlightcol} \relax
  [ \tif{t}{b. P}{u}{v} ] &\coloneqq&
  \pmatch{[t]}{b. \transl{P}}{\branch{\inl\ \tunit}{[u]} \\ \branch{\inr\ \tunit}{[v]}} \\
  \relax
  [ \inl\ t ] &\coloneqq& \inl\ [t] \\ \relax
  [ \inr\ t ] &\coloneqq& \inr\ [t] \\ \relax
  \left[ \pmatch{t}{p. P}{\branch{\inl\ u}{a} \\ \branch{\inr\ v}{b}} \right]
  &\coloneqq&
  \pmatch{[t]}{p. \transl{P}}{\branch{\inl\ u}{[a]} \\ \branch{\inr\ v}{[b]}} \\
  \relax
  [ \Ty{i} ] &\coloneqq& \Ty{i} \\ \relax
  [ \Pi (x : A).\ B ] &\coloneqq& \Pi (x : \transl{A}).\ \transl{B} \\
  \rowcolor{highlightcol} \relax
  [ \bool ] &\coloneqq& \unit + \unit \\ \relax
  \\
  \transl{A} &\coloneqq& [A]
\end{array}
\]

Once again this translation preserves typing, and is conservative (in particular
it preserves falsehood). So it offers a \emph{computational} evidence that
booleans are just a convenience provided you already have the \(\unit\) type and
sums. In that sense, a translation can---and perhaps should---be seen as
\emph{compilation} or as a compilation phase.
\marginnote[-2.3cm]{
  It is conservative because the translation behaves as the identity on the
  target.
}